---
title: "Brown bear RovQuant analysis"
subtitle: "OPSCR analysis workflow with the 'rovquantR' package"
author: "Pierre Dupont"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{Bear Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette reproduces the Open-Population Spatial Capture-Recapture analysis workflow from project RovQuant for the Brown bear population in Norway.
Specifically, we implement the latest version of the bear
model [@dupontBear2023] using nimble [@de2017programming;@nimbleSoftware2020] and nimbleSCR [@nimblescr2020bischof].  Details of the functions and procedure are provided therein.

This R script performs:
1. the initial cleaning of the large carnivore NGS data downloaded from RovBase.3.0
2. the data preparation for analysis with the 'nimbleSCR' package
3. the model fitting using 'nimble' and 'nimbleSCR'
4. the post-processing of the MCMC output



## Load Libraries 

```{r hidden, echo = FALSE, warning = FALSE, message = FALSE}
rm(list = ls())
gc()
```


```{r libraries, warning = FALSE, message = FALSE}
#library(rovquantR)
library(nimbleSCR)
```



## Define working environment

Here, we define the directories containing all the relevant data for the analysis and prepare a folder structure to store all the outputs. These include, among other things:
1. a clean version of the large carnivore NGS data for the species and period of interest, together with an .html report describing the content of the data.
2. OPSCR model inputs, one for each MCMC chain, containing the input data in the format required by 'nimbleSCR' and the OPSCR model 'nimble' definition, together with an .html report describing the OPSCR model (e.g. number of detections, habitat windows, inidviduals detected, covariates, ...)
3. directories to store model outputs such as density maps, tables of abundance and vital rates,...

```{r setup}
library(devtools)
#library(dplyr)
load_all("rovquantR")


##-- DATA DIRECTORY
##-- Directory containing the raw data necessary for the analysis
##-- (NB: This is NOT the working directory; NOTHING SHOULD BE SAVED/WRITTEN IN THIS DIRECTORY)
data_dir <- "C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/bear/2023/RovQuant_test/Data"


##-- WORKING DIRECTORY (= main folder for the analysis)
working_dir <- "C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/bear/2023/RovQuant_test/test1"


##-- Create folder structure for the analysis
makeDirectories(path = working_dir)
```



## Load and fix ROVBASE.3.0 data 

### 1. NGS data 

```{r }
##-- Load raw excel file imported from rovbase 
DNA <- readxl::read_xlsx(file.path(data_dir,"DNA.xlsx")) %>%
  ##-- Remove any duplicates
  distinct(., .keep_all = TRUE) %>%
  ##-- Rename columns to facilitate manipulation
  rename(., any_of(c( DNAID_sample = "DNAID (Prøve)",
                      Barcode_sample = "Strekkode (Prøve)",
                      RovbaseID_sample = "RovbaseID (Prøve)",
                      EventID = "HendelseID",
                      Analysis_priority = "Analyseprioritet",
                      Species_sample = "Art (Prøve)",
                      Sample_type = "Prøvetype", 
                      Date = "Funnetdato",
                      Mountain_area = "Fjellområde",
                      Sensitivity = "Følsomhet",
                      Release_Date = "Frigivelsesdato", 
                      Locality = "Lokalitet",
                      Location = "Funnsted",
                      North_original = "Nord (opprinnelig)",
                      East_Original = "Øst (opprinnelig)",
                      North_UTM33 = "Nord (UTM33/SWEREF99 TM)",
                      East_UTM33 = "Øst (UTM33/SWEREF99 TM)" ,
                      North_RT90 = "Nord (RT90)",
                      East_RT90 = "Øst (RT90)" ,
                      Coordinate_system = "Koordinatsystem" ,
                      Site_quality = "Stedkvalitet",
                      Collected_by = "Hvem samlet inn",
                      Collector_name = "Samlet selv - Navn",
                      Collector_phone = "Samlet selv - Telefon", 
                      Collector_email = "Samlet selv - E-post",
                      Collector_role = "Samlet selv - Rolle",
                      Collector_other_name = "Annen innsamler - Navn" ,
                      Collector_other_phone = "Annen innsamler - Telefon", 
                      Collector_other_email = "Annen innsamler - E-post",
                      Collector_other_role = "Annen innsamler - Rolle",
                      Tips_name = "Tipser - Navn",
                      Tips_phone = "Tipser - Telefon",
                      Tips_email = "Tipser - E-post",
                      Tips_role = "Tipser - Rolle",
                      Quality_checked = "Kvalitetssikret av feltpersonell",
                      Quality_check_name = "Kvalitetssikrer - navn",
                      Quality_check_orga = "Kvalitetssikrer - Organisasjon",
                      Comments_sample = "Merknad (Prøve)",
                      Last_saved_by_sample = "Sist lagret av (Prøve)",
                      Last_saved_sample = "Sist lagret dato (Prøve)",
                      DNAID = "DNAID (Analyse)",
                      Barcode = "Strekkode (Analyse)",
                      RovbaseID = "RovbaseID (Analyse)",
                      Species = "Art (Analyse)",
                      Sample_status = "Prøvestatus",
                      Id = "Individ",
                      Sex_analysis = "Kjønn (Analyse)",
                      Sex = "Kjønn (Individ)",
                      Method = "Metode",
                      Analyzed_by = "AnalysertAv",
                      Comments = "Merknad (Analyse)",
                      Last_saved_by = "Sist lagret av (Analyse)" ,
                      Last_saved = "Sist lagret dato (Analyse)" ,
                      Municipality_number = "Kommunenummer",
                      Municipality = "Kommune", 
                      County_number = "Fylkenummer",
                      County = "Fylke"))) %>%
  ##-- Filter to the focal species
  filter(., Species == "Bjørn") %>%
  ##-- Save as.csv 
  writeMostRecent.csv( ., file = file.path(data_dir, "dna_bear.csv"))

```

### 2. Dead recovery data

```{r}
##-- Load raw excel file imported from rovbase 
DR <- readxl::read_xlsx(file.path(data_dir,"dead carnivores.xlsx")) %>%
  ##-- Remove any duplicates
  distinct(., .keep_all = TRUE) %>%
  ##-- Rename columns to facilitate manipulation
  rename(., any_of(c( Species = "Art",
                      Death_cause = "Bakgrunn/årsak",
                      Death_method = "Bakgrunn/årsak metode",
                      Death_purpose = "Bakgrunn/årsak formål", 
                      Date = "Dødsdato",
                      Uncertain_date = "Usikker dødsdato",
                      Time_of_death = "Dødstidspunkt",
                      Age_class = "Alder på dødt individ",
                      Age_class_verif = "Aldersklasse verifisert SVA",
                      Juvenile = "Yngling",
                      Sex = "Kjønn", 
                      Age_estimated = "Alder, vurdert",
                      Age = "Alder, verifisert",
                      Age_verif_by = "Alder, verifisert av",
                      Full_weight =  "Helvekt",
                      Slaughter_weight =  "Slaktevekt",
                      CITES = "CITES-nummer",
                      Felling_site_verif = "Kontroll av fellingsted",
                      Tissue_sample = "Vevsprøve tatt",
                      Hunting_date = "Observasjons/Jaktdato",
                      Field_personnel ="Feltpersonell",
                      Control_status = "Kontrollstatus",
                      Assessment = "Vurdering",
                      Location = "Funnsted",
                      North_original = "Nord (opprinnelig)",
                      East_Original = "Øst (opprinnelig)",
                      North_UTM33 = "Nord (UTM33/SWEREF99 TM)",
                      East_UTM33 = "Øst (UTM33/SWEREF99 TM)",
                      North_RT90 = "Nord (RT90)",
                      East_RT90 = "Øst (RT90)",
                      Coordinate_system = "Koordinatsystem",
                      Site_quality = "Stedkvalitet",
                      Approved_date = "Godkjentdato",
                      Outcome = "Utfall", 
                      Counted_off_against_decision = "Regnes av mot vedtak", 
                      Approved_by = "Godkjent av",
                      Sensitivity = "Følsomhet",
                      Release_Date = "Frigivelsesdato", 
                      Id = "Individ", 
                      SVAID = "SVAID",
                      Lansstyrelsen_number = "Länsstyrelsens nr",
                      Last_saved_by = "Sist lagret av",
                      Last_saved =  "Sist lagret dato",
                      Municipality_number =  "Kommunenummer",
                      Municipality = "Kommune", 
                      County_number = "Fylkenummer",
                      County = "Fylke"))) %>%
  ##-- Filter to the focal species
  filter(., Species == "Bjørn") %>%
  ##-- Save as.csv 
  writeMostRecent.csv( ., file = file.path(data_dir, "dead_bear.csv"))

```



## Clean and check ROVBASE data

```{r, eval=FALSE}
cleanRovbaseData( species = "bear",
                  years = 2020:2024,
                  data_dir = data_dir,
                  output_dir = file.path(working_dir, "data"))
```



## Prepare OPSCR data in 'nimbleSCR' format

```{r,eval=FALSE}
makeRovquantData( species = "bear",
                  data_dir = data_dir,
                  working_dir = working_dir)


```



## Fit OPSCR model using 'nimble'

### 1. Females 

```{r, eval = FALSE}
##-- List all prepared input files
inputFiles <- list.files(file.path(working_dir, "nimbleInFiles/Hunn"),
                         full.names = T)

##-- Load the first one
load(inputFiles[1]) 

##-- Build nimble model object
model <- nimbleModel( code = modelCode,
                      constants = nimConstants,
                      inits = nimInits,
                      data = nimData,
                      check = FALSE,
                      calculate = FALSE) 
model$calculate()
cmodel <- compileNimble(model)
conf <- configureMCMC( model,
                       monitors = nimParams,
                       thin = 1,
                       monitors2 = nimParams2,
                       thin2 = 10)
Rmcmc <- buildMCMC(conf)
compiledList <- compileNimble( list(model = model,
                                    mcmc = Rmcmc),
                               showCompilerOutput = F)
Cmcmc <- compiledList$mcmc

##-- RUN NIMBLE MCMC IN SUCCESSIVE BITES
system.time(runMCMCbites( mcmc = Cmcmc,
                          bite.size = 200,
                          bite.number = 5,
                          path = file.path(working_dir,"nimbleOutfiles/Hunn")))


```

###  2. Males 

```{r, eval = FALSE}
##-- List all prepared input files
inputFiles <- list.files(file.path(working_dir, "nimbleInFiles/Hann"),
                         full.names = T)

##-- Load the first one
load(inputFiles[1]) 

##-- Build nimble model object
model <- nimbleModel( code = modelCode,
                      constants = nimConstants,
                      inits = nimInits,
                      data = nimData,
                      check = FALSE,
                      calculate = FALSE) 
model$calculate()
cmodel <- compileNimble(model)
conf <- configureMCMC(model,
                      monitors = nimParams,
                      thin = 1,
                      monitors2 = nimParams2,
                      thin2 = 10)
Rmcmc <- buildMCMC(conf)
compiledList <- compileNimble(list(model = model,
                                   mcmc = Rmcmc),
                              showCompilerOutput = F)
Cmcmc <- compiledList$mcmc

##-- RUN NIMBLE MCMC IN SUCCESSIVE BITES
system.time(runMCMCbites( mcmc = Cmcmc,
                          bite.size = 200,
                          bite.number = 5,
                          path = file.path(working_dir,"nimbleOutfiles/Hann")))



```



## Process OPSCR output

```{r, eval = FALSE}
processRovquantOutput( species = "bear",
                       data_dir = data_dir,
                       working_dir = working_dir,
                       extraction.raster = habitatRasterResolution$`5km`[["Regions"]])

```


## Results


## References


