---
title: "Bear Example"
subtitle : "RovQuant OPSCR analysis"
author: "Pierre Dupont"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{Wolverine Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette presents a workflow for the analysis of the Scandinavian Brown bear monitoring data as performed by the **[RovQuant project](https://www.nmbu.no/en/projects/rovquant)**. It consists in the Open-Population Spatial Capture-Recapture (OPSCR) analysis of the non-invasive genetic sampling and dead recovery data collected by the Norwegian and Swedish authorities and made available in the transnational large carnivore database **[Rovbase 3.0](https://rovbase.no/)**.

Since 2017, the **Applied Quantitative Ecology Group** ([AQEG](https://www.nmbu.no/en/research/groups/applied-quantitative-ecology-group-aqeg)) has developed specialized methods for the Spatial Capture-Recapture analysis of large-scale individual-based datasets in a Bayesian framework.

These developments were conducted in collaboration with the **[NIMBLE](https://r-nimble.org/)** development team, and led to the publication of the **[nimbleSCR](https://cran.r-project.org/web/packages/nimbleSCR/index.html)** R package compiling a set of custom functions and distributions. 

```{r, echo = FALSE}
rm(list = ls())
gc()
```



# Data Preparation

We start by loading the different R packages necessary for the analysis, and in particular the 'rovquantR' package, which contains a set of wrapper functions to facilitate the analysis. 

```{r library, warning = FALSE, message = FALSE}
library(devtools)
library(dplyr)
library(readxl)
library(nimbleSCR)
library(rovquantR)
```



## Working Environment

Next, we need to set-up the general working environment for the analysis. This includes the location of the raw data files (imported from RovBase) and the directory chosen for the analysis. 
The first directory is called 'data_dir' and corresponds to the folder where the raw data is located. 

It contains the different files necessary for the OPSCR analysis:
* an excel file of all the brown bear DNA samples available from Rovbase
* an excel file of all the brown bear dead recoveries available from Rovbase
* an excel file of all large carnivore observations available from Skandobs
* multiple spatial covariates (in the 'GIS' folder)

Note that this is NOT the working directory. For all intent and purposes, it should be treated as a 'read-only' directory.

```{r set-up1.1, eval = TRUE, echo = FALSE}
data_dir <- "C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/bear/2023/RovQuant_test/Data"
```

```{r set-up1.2, eval = FALSE, echo = TRUE}
data_dir <- "./Data"
```


The second directory will be used to store all the files and outputs from the analysis.

```{r set-up2.1, eval = TRUE, echo = FALSE}
working_dir <- "C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/bear/2023/RovQuant_test/BearVignette"
``` 

```{r set-up2.2, eval = FALSE, echo = TRUE}
working_dir <- "./BearVignette"
``` 


To make it easier (and more reproducible), we created the 'makeDirectories' function. This will create a set of folders in the location specific by 'working_dir' that will later be used to store processed data, model input and output, as well as results figures and tables.

```{r set.up3, eval = TRUE}
makeDirectories(path = working_dir)
``` 



## Rovbase data

Before we dive in the OPSCR analysis itself, there is one more step we need to take.
The raw export from Rovbase contain variable column names and Scandinavian characters which make it cumbersome to work with in R. We therefore rename the columns here for easier processing.

In addition, we have noted that, over the years, certain fields have been added, while others had their names changed. For this reason, it is very difficult/impossible to automatize this process and the user has to make sure the field names below are up to date.

```{r rename_DNA, eval=FALSE}
##-- Load raw excel file imported from rovbase 
DNA <- readxl::read_xlsx(file.path(data_dir,"DNA.xlsx")) %>%
  ##-- Remove any duplicates
  distinct(., .keep_all = TRUE) %>%
  ##-- Rename columns to facilitate manipulation
  rename(., any_of(c( DNAID_sample = "DNAID (Prøve)",
                      Barcode_sample = "Strekkode (Prøve)",
                      RovbaseID_sample = "RovbaseID (Prøve)",
                      EventID = "HendelseID",
                      Analysis_priority = "Analyseprioritet",
                      Species_sample = "Art (Prøve)",
                      Sample_type = "Prøvetype", 
                      Date = "Funnetdato",
                      Mountain_area = "Fjellområde",
                      Sensitivity = "Følsomhet",
                      Release_Date = "Frigivelsesdato", 
                      Locality = "Lokalitet",
                      Location = "Funnsted",
                      North_original = "Nord (opprinnelig)",
                      East_Original = "Øst (opprinnelig)",
                      North_UTM33 = "Nord (UTM33/SWEREF99 TM)",
                      East_UTM33 = "Øst (UTM33/SWEREF99 TM)" ,
                      North_RT90 = "Nord (RT90)",
                      East_RT90 = "Øst (RT90)" ,
                      Coordinate_system = "Koordinatsystem" ,
                      Site_quality = "Stedkvalitet",
                      Collected_by = "Hvem samlet inn",
                      Collector_name = "Samlet selv - Navn",
                      Collector_phone = "Samlet selv - Telefon", 
                      Collector_email = "Samlet selv - E-post",
                      Collector_role = "Samlet selv - Rolle",
                      Collector_other_name = "Annen innsamler - Navn" ,
                      Collector_other_phone = "Annen innsamler - Telefon", 
                      Collector_other_email = "Annen innsamler - E-post",
                      Collector_other_role = "Annen innsamler - Rolle",
                      Tips_name = "Tipser - Navn",
                      Tips_phone = "Tipser - Telefon",
                      Tips_email = "Tipser - E-post",
                      Tips_role = "Tipser - Rolle",
                      Quality_checked = "Kvalitetssikret av feltpersonell",
                      Quality_check_name = "Kvalitetssikrer - navn",
                      Quality_check_orga = "Kvalitetssikrer - Organisasjon",
                      Comments_sample = "Merknad (Prøve)",
                      Last_saved_by_sample = "Sist lagret av (Prøve)",
                      Last_saved_sample = "Sist lagret dato (Prøve)",
                      DNAID = "DNAID (Analyse)",
                      Barcode = "Strekkode (Analyse)",
                      RovbaseID = "RovbaseID (Analyse)",
                      Species = "Art (Analyse)",
                      Sample_status = "Prøvestatus",
                      Id = "Individ",
                      Sex_analysis = "Kjønn (Analyse)",
                      Sex = "Kjønn (Individ)",
                      Method = "Metode",
                      Analyzed_by = "AnalysertAv",
                      Comments = "Merknad (Analyse)",
                      Last_saved_by = "Sist lagret av (Analyse)" ,
                      Last_saved = "Sist lagret dato (Analyse)" ,
                      Municipality_number = "Kommunenummer",
                      Municipality = "Kommune", 
                      County_number = "Fylkenummer",
                      County = "Fylke"))) %>%
  ##-- Filter to the focal species
  filter(., Species == "Bjørn") %>%
  ##-- Save as.csv 
  writeMostRecent.csv( ., file = file.path(data_dir, "dna_bear.csv"))
```


We also have to rename the dead recovery file for easier handling.

```{r rename_DR, eval=FALSE}
##-- Load raw excel file imported from rovbase 
DR <- readxl::read_xlsx(file.path(data_dir,"dead carnivores.xlsx")) %>%
  ##-- Remove any duplicates
  distinct(., .keep_all = TRUE) %>%
  ##-- Rename columns to facilitate manipulation
  rename(., any_of(c( Species = "Art",
                      Death_cause = "Bakgrunn/årsak",
                      Death_method = "Bakgrunn/årsak metode",
                      Death_purpose = "Bakgrunn/årsak formål", 
                      Date = "Dødsdato",
                      Uncertain_date = "Usikker dødsdato",
                      Time_of_death = "Dødstidspunkt",
                      Age_class = "Alder på dødt individ",
                      Age_class_verif = "Aldersklasse verifisert SVA",
                      Juvenile = "Yngling",
                      Sex = "Kjønn", 
                      Age_estimated = "Alder, vurdert",
                      Age = "Alder, verifisert",
                      Age_verif_by = "Alder, verifisert av",
                      Full_weight =  "Helvekt",
                      Slaughter_weight =  "Slaktevekt",
                      CITES = "CITES-nummer",
                      Felling_site_verif = "Kontroll av fellingsted",
                      Tissue_sample = "Vevsprøve tatt",
                      Hunting_date = "Observasjons/Jaktdato",
                      Field_personnel ="Feltpersonell",
                      Control_status = "Kontrollstatus",
                      Assessment = "Vurdering",
                      Location = "Funnsted",
                      North_original = "Nord (opprinnelig)",
                      East_Original = "Øst (opprinnelig)",
                      North_UTM33 = "Nord (UTM33/SWEREF99 TM)",
                      East_UTM33 = "Øst (UTM33/SWEREF99 TM)",
                      North_RT90 = "Nord (RT90)",
                      East_RT90 = "Øst (RT90)",
                      Coordinate_system = "Koordinatsystem",
                      Site_quality = "Stedkvalitet",
                      Approved_date = "Godkjentdato",
                      Outcome = "Utfall", 
                      Counted_off_against_decision = "Regnes av mot vedtak", 
                      Approved_by = "Godkjent av",
                      Sensitivity = "Følsomhet",
                      Release_Date = "Frigivelsesdato", 
                      Id = "Individ", 
                      SVAID = "SVAID",
                      Lansstyrelsen_number = "Länsstyrelsens nr",
                      Last_saved_by = "Sist lagret av",
                      Last_saved =  "Sist lagret dato",
                      Municipality_number =  "Kommunenummer",
                      Municipality = "Kommune", 
                      County_number = "Fylkenummer",
                      County = "Fylke"))) %>%
  ##-- Filter to the focal species
  filter(., Species == "Bjørn") %>%
  ##-- Save as.csv 
  writeMostRecent.csv( ., file = file.path(data_dir, "dead_bear.csv"))
```



## Data cleaning

Now, we clean the data. This step involves multiples actions, such as subsetting the data to the species and period of interest for the analysis,  checking individual sex assigments, etc...
In order to evaluate the cleaning procedure, the cleaing function prints out a report in .html format summarizing the content of the cleaned dataset. This report can be found in the 'data' folder of the analysis directory, labbled with the date corresponding to the last modification date of the raw data.

```{r clean}
cleanRovbaseData( species = "bear",
                  years = 2020:2024,
                  data_dir = data_dir,
                  output_dir = file.path(working_dir, "data"))
```



## Data preparation

Now that we have cleaned out the data, it is time to format it for analysis using nimbleSCR.
This process can be quite opaque and complex and this is why we have created the 'makeRovquantData' wrapper function, which encapsulates the latest data formatting scripts for the different species.

```{r make}
makeRovquantData(    
  ##-- paths
  species = "bear",
  data_dir = data_dir,
  working_dir = working_dir,
  ##-- data
  years = NULL,
  sex = c("Hann","Hunn"),
  aug.factor = 2,
  sampling.months = list(4,5,6,7,8,9,10,11),
  ##-- habitat
  habitat.res = 20000, 
  buffer.size = 50000,
  max.move.dist = 250000,
  ##-- detectors
  detector.res = 5000,
  subdetector.res = 1000,
  max.det.dist = 70000,
  resize.factor = 1,
  ##-- miscellanious
  plot.check = FALSE,
  print.report = TRUE)
```



# Model fitting

The next step is to fit the OPSCR model to the prepared bear data using NIMBLE.
In this example, as in most cases in the RovQuant project, we fit sex-specific models.

```{r run_females1.1, eval=FALSE}
##-- List all prepared input files
inputFiles <- list.files(file.path(working_dir, "nimbleInFiles/Hunn"),
                         full.names = T)

##-- Load the first one
load(inputFiles[1]) 
```

```{r run_females1.2, eval=FALSE}
 load(system.file("extdata", "WolverineData.RData", package = "nimbleSCR"))
```  


The first step when fitting a statistical model using NIMBLE is to create a model object, containing the model code, the constants/dimensions, data and initial values used in the model.




```{r run_females2, eval=FALSE}
model <- nimbleModel( code = modelCode,
                      constants = nimConstants,
                      inits = nimInits,
                      data = nimData,
                      check = FALSE,
                      calculate = FALSE) 
```


The second step is to compile the model in C++.

```{r run_females3, eval=FALSE}
cmodel <- compileNimble(model)
```


Next, we need to configure and compile the MCMC algorithm that will be used to sample posterior samples.

```{r run_females4, eval=FALSE}
conf <- configureMCMC( model,
                       monitors = nimParams,
                       thin = 1,
                       monitors2 = nimParams2,
                       thin2 = 5)
Rmcmc <- buildMCMC(conf)
compiledList <- compileNimble( list(model = model,
                                    mcmc = Rmcmc),
                               showCompilerOutput = F)
Cmcmc <- compiledList$mcmc
```


Finally, once the MCMC algorithm and the model are prepared and compiled, we can run it using the 'runMCMCbites' function.

```{r run_females5, eval=FALSE}
system.time(runMCMCbites( mcmc = Cmcmc,
                          bite.size = 100,
                          bite.number = 5,
                          path = file.path(working_dir,"nimbleOutfiles/Hunn")))

```


And now we can do the same for males:
```{r run_males, eval=FALSE}
##-- Load the first prepared input files
inputFiles <- load(list.files(file.path(working_dir, "nimbleInFiles/Hann"),
                              full.names = T)[1])

##-- Build nimble model object
model <- nimbleModel( code = modelCode,
                      constants = nimConstants,
                      inits = nimInits,
                      data = nimData,
                      check = FALSE,
                      calculate = FALSE) 
model$calculate()
cmodel <- compileNimble(model)
conf <- configureMCMC(model,
                      monitors = nimParams,
                      thin = 1,
                      monitors2 = nimParams2,
                      thin2 = 5)
Rmcmc <- buildMCMC(conf)
compiledList <- compileNimble(list(model = model,
                                   mcmc = Rmcmc),
                              showCompilerOutput = F)
Cmcmc <- compiledList$mcmc

##-- Run nimble MCMC in successive bites
system.time(runMCMCbites( mcmc = Cmcmc,
                          bite.size = 100,
                          bite.number = 5,
                          path = file.path(working_dir,"nimbleOutfiles/Hann")))
```



# Output processing



```{r process, eval=FALSE}
processRovquantOutput(   
  ##-- paths
  species = "bear",
  data_dir = data_dir,
  working_dir = working_dir,
  ##-- MCMC processing
  nburnin = 0,
  ##-- Density extraction
  niter = 100,
  extraction.res = 5000,
  ##-- miscellanious
  plot.check = FALSE,
  print.report = TRUE)
```



```{r visualize}

```