---
title: "rovquantR - Wolverine debugging"
author: "Pierre Dupont"
format: html
editor: visual
---

## Introduction

This document presents the process of tracking down the differences between the Wolverine data considered when using the **rovquantR** package and last year's analysis.

## Set-up Working Environment

```{r}
rm(list=ls())
gc()

library(rovquantR)
library(sf)
library(raster)
library(dplyr)
```

First, we need to set-up some basic variables:

```{r}
years <- 2015:2024
sampling.months <- list(12,1:6)
yearsSampledNorrb <- c(2016:2018,2023,2024)
```

Next, we can load a shapefile of the **COUNTIES** used for the sample assignment. Note that the current **rovquantR** package uses a slightly different shapefile for the **COUNTIES** leading to marginally different samples designation to counties/countries (+ or - 2-3 samples). This is tolerable and not the problem we are exploring here.

```{r}
source("C:/My_documents/RovQuant/Temp/PD/myWorkingDirectories.R")
source("C:/My_documents/RovQuant/Source/CleanDataNew2sf.R")
source("C:/My_documents/RovQuant/Source/FilterDatasf.R")
source("C:/My_documents/RovQuant/Source/MakeHabitatFromRastersf.R")

load(file.path(dir.dropbox,"/DATA/MISC DATA/age.lookup.table.RData"))

COMMUNES_NOR <- st_read(file.path(dir.dropbox, "DATA/GISData/scandinavian_border/NOR_adm2_UTM33.shp"))
COMMUNES_SWE <- st_read(file.path(dir.dropbox, "DATA/GISData/scandinavian_border/SWE_adm2_UTM33.shp"))
COMMUNES <- rbind(COMMUNES_NOR, COMMUNES_SWE)
COUNTIES <- COMMUNES %>%
  group_by(NAME_1) %>%
  summarize()
COUNTIESNorrbotten <- COUNTIES[COUNTIES$NAME_1 %in% "Norrbotten", ]
```

## Filtered Data Comparison

### **rovquantR** data

Now we can load the data cleaned by the **rovquantR** package. At this stage, the data still contains :

1.  samples in Norrbotten in years without sampling

2.  samples outside the years considered for this analysis

3.  samples outside the monitoring period

```{r}
working.dir <- "C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/wolverine/2025/Test_rovquantR"

myFullData.sp <- readMostRecent( 
  path = file.path(working.dir,"data"),
  pattern = "CleanData_wolverine",
  extension = ".RData") 

dim(myFullData.sp$alive)
```

Now, we can load the data cleaned by the **rovquantR** package. At this stage, the data still contains :

1.  samples outside Norway and Sweden

2.  samples in Norrbotten in years without sampling

3.  samples outside the years considered for this analysis

4.  samples outside the monitoring period

5.  samples with unknown sex

```{r}
##-- Filter NGS samples
myFullData.sp$alive <- myFullData.sp$alive %>%
  dplyr::filter(
    ##-- Subset to years of interest
    Year %in% years,
    ##-- Subset to monitoring period
    Month %in% unlist(sampling.months), 
    ##-- Remove samples with unknown sex
    Sex %in% c("male","female"),
    ##-- Subset to samples collected in Norway and Sweden
    Country_sf %in% c("(N)","(S)"))
    #Country_sample %in% c("(N)","(S)"))
dim(myFullData.sp$alive)
```

Here too, the way to sub-sample the data can lead to marginally different results (but within the tolerance margin). If it is based on the country registered in **RovBase** ('**Country_sample**'), 14 more samples are kept compared to when using a filter based on the overlap with the **COUNTRY** shapefile ('**Country_sf**').

We also need to remove samples in Norrbotten in years without systematic sampling:

```{r}
##-- Identify samples removed 
yearsNotSampled <- years[!years %in% yearsSampledNorrb]
##-- Identify detections collected in Norrbotten 
is.Norr2 <- as.numeric(st_intersects(myFullData.sp$alive, COUNTIESNorrbotten))
sum(is.Norr2, na.rm = T)
table(myFullData.sp$alive$Year[(myFullData.sp$alive$Year %in% yearsNotSampled & is.Norr2 %in% 1)])

##-- Filter out detections in Norrbotten in years without sampling
test2 <- myFullData.sp$alive %>%
  dplyr::filter(!(Year %in% yearsNotSampled & is.Norr2 %in% 1))
dim(test2)

##-- Identify detections still in Norrbotten 
is.Norr2 <- as.numeric(st_intersects(test2, COUNTIESNorrbotten))
sum(is.Norr2, na.rm = T)
table(test2$Year[is.Norr2 %in% 1])
table(test2$Year[is.na(is.Norr2)])
table(test2$Year)
```

### '**54.Cleaned2025**' data

We can now load the filtered data for last year's analysis ('**54.Cleaned2025**') for comparison:

```{r}
load("C:/Users/pidu/AQEG Dropbox/AQEG Team Folder/RovQuant/wolverine/CM/2025/54.Cleaned2025/myFilteredData_original.RData")
dim(myFilteredData.sp$alive)
dim(myFilteredData.sp$dead.recovery)
extent(myFilteredData.sp$alive)
is.Norr <- as.numeric(st_intersects(
  myFilteredData.sp$alive,
  COUNTIESNorrbotten))
sum(is.Norr, na.rm = T)
table(myFilteredData.sp$alive$Year[!is.na(is.Norr)])
table(myFilteredData.sp$alive$Year[is.na(is.Norr)])
table(myFilteredData.sp$alive$Year)
```

### Data differences

We see that differences in the number of NGS samples retained are spread across years and within/outside Norrbotten:

```{r}
##-- Differences in data after filtering 

##-- Total number of NGS samples
nrow(test2) - nrow(myFilteredData.sp$alive)

##-- Number of samples kept in Norrbotten
sum(is.Norr2, na.rm = T) - sum(is.Norr,na.rm=T)

##-- Number of samples in Norrbotten (kept) per year
table(test2$Year[is.Norr2 %in% 1])-table(myFilteredData.sp$alive$Year[!is.na(is.Norr)]) 

##-- Number of samples outside of Norrbotten
table(test2$Year[is.na(is.Norr2)])-table(myFilteredData.sp$alive$Year[is.na(is.Norr)])

##-- Number of samples overall
table(test2$Year)-table(myFilteredData.sp$alive$Year)
```

## Debugging

### Data cleaning and filtering

In this section, we will track down the additional samples kept by the **rovquantR** package and try to identify when they are removed during the data processing of last year.

#### Problematic samples

We start by listing all problematic samples:

```{r}
##-- Identify how many samples are in the NEW but not OLD data
sum(!test2$DNAID %in% myFilteredData.sp$alive$DNAID)

##-- Store samples that are in the NEW but not OLD data
Probs <- test2[!test2$DNAID %in% myFilteredData.sp$alive$DNAID, ]
table(Probs$Year)
```

#### Load raw data

Then, we will copy-paste and go through the different steps from last year's script ('**54.Cleaned2025TestPDScript.R**') and check when and why these additional samples are removed.

The first step is to load the raw files exported from Rovbase and used in last year's analysis. The first one is the NGS file.

*At this point,* *all "problematic" samples are still in the NGS dataset*:

```{r}
DNA <- read.csv( 
  file.path(dir.dropbox, "DATA/RovbaseData/ROVBASE DOWNLOAD 20251121/dna_wolverines.csv"),
  fileEncoding = "latin1") 
all(Probs$DNAID %in% DNA$DNAID..Prøve.)
```

We can also check that these samples are not present in the dead recovery data:

```{r}
DEAD <- read.csv( 
  file.path(dir.dropbox, "DATA/RovbaseData/ROVBASE DOWNLOAD 20251121/dead_carnivores.csv"),
  fileEncoding = "latin1") 
sum(Probs$DNAID %in% DEAD$DNAID)
```

We can also check that none of these "problematic" samples are in the suspect samples flagged by Henrik Brøseth:

```{r}
## DNA samples to be removed from Henrik (no samples to be removed from Sweden)
SUSPECT_NGS_SAMPLES <- read.csv( file.path(dir.dropbox, "DATA/RovbaseData/ROVBASE DOWNLOAD 20241023/Remove ngs samples list wolverine 2024.csv"),
                                 fileEncoding = "latin1") 
SUSPECT_NGS_SAMPLES_2025 <- read.csv( file.path(dir.dropbox, "DATA/RovbaseData/ROVBASE DOWNLOAD 20251121/Additional samples to add to the Remove ngs samples list wolverine from Norway 2025.csv"),
                                      fileEncoding = "latin1")
SUSPECT_NGS_SAMPLES <- rbind(SUSPECT_NGS_SAMPLES, SUSPECT_NGS_SAMPLES_2025)  # rbind file from last year with the one from 2025
sum(Probs$DNAID %in% SUSPECT_NGS_SAMPLES$DNAID_RB)

## DEAD RECOVERIES to be removed from Henrik  --> Only file from 2024, NO ADDITIONAL INDIVIDUALS IN 2025.
SUSPECT_DeadRecoSAMPLES <- read.csv( file.path(dir.dropbox, "DATA/RovbaseData/ROVBASE DOWNLOAD 20241023/Remove dead recoveries list wolverine 2024.csv"),
                                     fileEncoding = "latin1") 
sum(Probs$RovbaseID %in% SUSPECT_DeadRecoSAMPLES$Rovbase_ID)
```

Then, we translate column names, and remove potential dead recoveries from the NGS file.

*All "problematic" samples are still in the NGS dataset*.

```{r}
##-- TRANSLATE SCANDINAVIAN CHARACTERS 
colnames(DNA) <- translateForeignCharacters( 
  dat = colnames(DNA))
## Drop a column that makes cleanDataNew to fail
DNA <- DNA[ ,-which(colnames(DNA)%in% "Kjoenn..Individ.")]
colnames(DEAD) <- translateForeignCharacters(
  dat = colnames(DEAD))
all(Probs$DNAID %in% DNA$DNAID..Proeve.)

##-- CLEAN & FILTER NGS DATA
## Remove DEAD entries from the DNA data [HB]
DNA <- DNA[substr(DNA$RovbaseID..Proeve.,1,1) != "M", ]
all(Probs$DNAID %in% DNA$DNAID..Proeve.)

##-- Remove un-verified dead recoveries [HB]
## ("Påskutt ikke belastet kvote" & "Påskutt belastet kvote")
DEAD <- DEAD[!grepl(pattern = "Påskutt", x = as.character(DEAD$Utfall)), ]
any(Probs$DNAID %in% DEAD$DNAID)
```

#### CleanDataNew( )

The next step is the '**CleanDataNew2sf( )**' function.

*All "problematic" samples are still in the NGS dataset*.

```{r}
##-- CLEAN NGS & DEAD RECOVERY DATA 
myCleanedData.sp <- CleanDataNew2sf( 
  dna_samples = DNA,
  dead_recoveries = DEAD,
  species_id = "Jerv",
  country_polygon = COUNTRIES,
  threshold_month = unlist(sampling.months)[1],
  keep_dead = T,
  age.label.lookup = age.lookup.table)

sum(Probs$DNAID %in% myCleanedData.sp$DNAID)
```

#### FilterData( )

The next step is the **'FilterDatasf( )'** function. First, we try with the **COUNTRY** shapefile provided in the **rovquantR** package (i.e. without lakes).

**This DOES NOT remove any "problematic" samples from the NGS data !!!!**

```{r}
data("COUNTRIES")

myStudyArea_NEW <- COUNTRIES[COUNTRIES$ISO %in% c("SWE","NOR"), ]
myStudyArea_NEW$id <- 1
myStudyArea_NEW <- myStudyArea_NEW %>% group_by(id) %>% summarize()


### ====    1.2. FILTER DATA ====

myFullData_NEW <- FilterDatasf(
  myData = myCleanedData.sp,
  poly = myStudyArea_NEW,
  dead.recovery = T ,
  sex = c("Hann","Hunn"), 
  setSex = T)

dim(myFullData_NEW$alive)
table(myFullData_NEW$alive$Year)
sum(Probs$DNAID %in% myFullData_NEW$alive$DNAID)
any(Probs$DNAID %in% myFullData_NEW$dead.recovery$DNAID)
```

Next, we try with the old **COUNTRY** shapefiles (with lakes).

**This removes 125 "problematic" samples from the NGS data !!!!**

```{r}
##-- POLYGONS OF THE REGION 
GLOBALMAP <- st_read(file.path( dir.dropbox, 
                                "DATA/GISData/vegetation/Countries_waterHumans25000000m2_multimulti.shp"))  
GLOBALMAP <- GLOBALMAP[GLOBALMAP$area > 80000000, ] 
GLOBALMAP <- st_crop(GLOBALMAP, st_bbox(extent(c(-70000,1200000,5100000,8080000)))) 

##-- POLYGONS OF SWEDEN & NORWAY 
COUNTRIES <- GLOBALMAP[GLOBALMAP$ISO %in% c("SWE","NOR"), ] 
COUNTRIES <- COUNTRIES %>% 
  group_by(ISO) %>% 
  summarize() 

myStudyArea <- COUNTRIES[COUNTRIES$ISO %in% c("SWE","NOR"), ] 
myStudyArea$id <- 1 
myStudyArea <- myStudyArea %>% group_by(id) %>% summarize()   


### ====    1.2. FILTER DATA ====  

myFullData.sp <- FilterDatasf( 
  myData = myCleanedData.sp,
  poly = myStudyArea,
  dead.recovery = T,
  sex = c("Hann","Hunn"),
  setSex = T)  

dim(myFullData.sp$alive) 
table(myFullData.sp$alive$Year)  
sum(Probs$DNAID %in% myFullData.sp$alive$DNAID) 
any(Probs$DNAID %in% myFullData.sp$dead.recovery$DNAID)  
```

#### Flagged samples

Next, we remove NGS samples and dead recoveries flagged by Henrik.

*No "problematic" samples are removed at this step.*

```{r}
## REMOVE SUSPECT SAMPLES ACCORDING TO HENRIK (THERE ARE NOT SUSPECT SAMPLES IN SWEDEN)
myFullData.sp$alive$DNAID <- as.character(myFullData.sp$alive$DNAID)
myFullData.sp$alive <- myFullData.sp$alive[!(myFullData.sp$alive$DNAID %in% as.character(SUSPECT_NGS_SAMPLES$DNAID_RB)), ]
dim(myFullData.sp$alive)

## REMOVE SUSPECT DEAD RECOVERIES ACCORDING TO HENRIK
myFullData.sp$dead.recovery$DNAID <- as.character(myFullData.sp$dead.recovery$DNAID)
myFullData.sp$dead.recovery <- myFullData.sp$dead.recovery[!(myFullData.sp$dead.recovery$RovBaseId %in% as.character(SUSPECT_DeadRecoSAMPLES$Rovbase_ID)), ]
dim(myFullData.sp$dead.recovery)
```

At this point, we remove manually three pup samples flagged by Eva Hedmark in an email.

**This removes the last 3 "problematic" samples in 'myFullData.sp\$alive'!!!!!**

```{r}
## REMOVE CUBS OF THE YEAR ACCORDING TO EVA HEDMARK (email 23/11/2025, document with explanation in AQEG Dropbox\AQEG Team Folder\RovQuant\DATA\RovbaseData\ROVBASE DOWNLOAD 20251121)
myFullData.sp$alive <- myFullData.sp$alive[!(myFullData.sp$alive$DNAID %in% c("D608410", "D608411", "D605997")),]

any(Probs$DNAID %in% myFullData.sp$alive$DNAID)
sum(Probs$DNAID %in% myFullData.sp$alive$DNAID)
```

Following are several steps to remove individuals with multiple dead recoveries. This does not change anything.

```{r}
## Remove individuals that died twice
myFullData.sp$dead.recovery$Id <- as.character(myFullData.sp$dead.recovery$Id)
IdDoubleDead <- myFullData.sp$dead.recovery$Id[duplicated(myFullData.sp$dead.recovery$Id)]

if(length(IdDoubleDead) > 0){
  duplicatedDeath <- NULL
  for(i in IdDoubleDead){
    tmp  <- which(myFullData.sp$dead.recovery$Id == i & is.na(myFullData.sp$dead.recovery$DeathCause_2))
    if(length(tmp)==0){tmp  <- which(myFullData.sp$dead.recovery$Id == i)[-2]}##[CM] remove the second record.
    duplicatedDeath <- c(duplicatedDeath, tmp)
  }#i
  myFullData.sp$dead.recovery <- myFullData.sp$dead.recovery[-duplicatedDeath, ]
}#if

## Dead recoveries flagged by Henrik that should always be removed (email from the 18/12/2024)
myFullData.sp$dead.recovery <- myFullData.sp$dead.recovery[!myFullData.sp$dead.recovery$RovBaseId %in% c("M495994", "M524051", "M524052", "M524053"),]


## Remove pups killed before recruitment based on weight (cf. Henrik)
## 1) Remove individuals that are "Ja" in column "Doedt.individ..Unge" and recovered dead between March and November
sum(myFullData.sp$dead.recovery$Alder.pa.doedt.individ %in% "Unge" &
      myFullData.sp$dead.recovery$Month > 2 &
      myFullData.sp$dead.recovery$Month < 12)

myFullData.sp$dead.recovery <- myFullData.sp$dead.recovery[-which(myFullData.sp$dead.recovery$Alder.pa.doedt.individ %in% "Unge" &
                                                                    myFullData.sp$dead.recovery$Month > 2 &
                                                                    myFullData.sp$dead.recovery$Month < 12),]


## 2) Remove individuals that have a weight >0 and <4 between March and November format the weight correctly
myFullData.sp$dead.recovery$Helvekt <- as.character(myFullData.sp$dead.recovery$Helvekt)
myFullData.sp$dead.recovery$Slaktevekt <- as.character(myFullData.sp$dead.recovery$Slaktevekt)

##-- Convert to decimals
myFullData.sp$dead.recovery$Helvekt <- as.numeric(gsub(",", ".", myFullData.sp$dead.recovery$Helvekt))
myFullData.sp$dead.recovery$Slaktevekt <- as.numeric(gsub(",", ".", myFullData.sp$dead.recovery$Slaktevekt))

##-- Get the two weight columns together.
myFullData.sp$dead.recovery$weight <- ifelse(!is.na(myFullData.sp$dead.recovery$Helvekt),
                                             myFullData.sp$dead.recovery$Helvekt,
                                             myFullData.sp$dead.recovery$Slaktevekt)

##-- Assign negative values to nas to avoid issues
myFullData.sp$dead.recovery$weight[is.na(myFullData.sp$dead.recovery$weight)] <- -999
dim(myFullData.sp$dead.recovery)

##-- check how many dead reco we remove and remove if more than 0
if(sum(myFullData.sp$dead.recovery$weight > 0 &
       myFullData.sp$dead.recovery$weight < 4 &
       myFullData.sp$dead.recovery$Month < 12 &
       myFullData.sp$dead.recovery$Month > 2)>0){
  myFullData.sp$dead.recovery <- myFullData.sp$dead.recovery[-which(myFullData.sp$dead.recovery$weight > 0 &
                                                                      myFullData.sp$dead.recovery$weight < 4 &
                                                                      myFullData.sp$dead.recovery$Month < 12 &
                                                                      myFullData.sp$dead.recovery$Month > 2),]
}

##-- check how many dead reco with a weight of 0 kg and recovered between march and november
if(sum(myFullData.sp$dead.recovery$Age %in% 0 &
       myFullData.sp$dead.recovery$Month < 12 &
       myFullData.sp$dead.recovery$Month > 2)>0){
  myFullData.sp$dead.recovery[myFullData.sp$dead.recovery$Age %in% 0 &
                                myFullData.sp$dead.recovery$Month < 12 &
                                myFullData.sp$dead.recovery$Month > 2,  ]
}

sum(Probs$DNAID %in% myFullData.sp$dead.recovery$DNAID)
```

The next step is removing samples outside of the monitoring period and years of interest.

```{r}
### ====    1.3. FILTER NGS & DEAD RECOVERY DATA ====

myFilteredData.sp <- myFullData.sp

## Subset to years of interest
myFilteredData.sp$alive <- myFilteredData.sp$alive[myFilteredData.sp$alive$Year %in% years, ]
sum(Probs$DNAID %in% myFilteredData.sp$alive$DNAID)

myFilteredData.sp$dead.recovery <- myFilteredData.sp$dead.recovery[myFilteredData.sp$dead.recovery$Year %in% years, ]
sum(Probs$DNAID %in% myFilteredData.sp$dead.recovery$DNAID)

## Subset to months of interest
myFilteredData.sp$alive <- myFilteredData.sp$alive[myFilteredData.sp$alive$Month %in% unlist(sampling.months), ]
sum(Probs$DNAID %in% myFilteredData.sp$alive$DNAID)

dim(myFilteredData.sp$alive)
table(myFilteredData.sp$alive$Year)
```

#### Samples in Norrbotten

The next step is removing samples in Norrbotten in years without systematic sampling.

*All remaining "problematic" samples are still in the NGS dataset*.

```{r}
COUNTIESNorrbotten <- COUNTIES[COUNTIES$NAME_1 %in% "Norrbotten", ]
yearsSampledNorrb <- c(2016:2018,2023,2024)
is.Norr <- as.numeric(st_intersects(myFilteredData.sp$alive, COUNTIESNorrbotten))

## Remove samples in Norrbotten in years not sampled
myFilteredData.sp$alive <- myFilteredData.sp$alive[- which(!myFilteredData.sp$alive$Year %in% yearsSampledNorrb & !is.na(is.Norr)), ]

dim(myFilteredData.sp$alive)
table(myFilteredData.sp$alive$Year)
sum(Probs$DNAID %in% myFilteredData.sp$alive$DNAID)
```

Finally we check the dimensions and distribution of samples of the saved NGS dataset to make sure they correspond:

```{r}
load(file.path(dir.dropbox,
               "wolverine/CM/2025/54.Cleaned2025/myFilteredData_original.RData"))

dim(myFilteredData.sp$alive)
table(myFilteredData.sp$alive$Year)
```

#### Conclusion

The discrepancy comes from the shapefile used to filter out samples outside of Norway/Sweden. In the old version, **125 NGS sample**s fall into lakes (within Norway or Sweden) and are thus removed.

In addition, **3 NGS samples** are removed manually after being flagged by Eva Hedmark.

We need to :

1.  check which shapefile to use (should we keep removing samples falling into lakes?)

2.  add the removal of pup NGS samples to the **makeRovquantData_wolverine( )** function.

## Generate Habitat

Here, we compare the habitat characteristics if they are generated with or without the **128 NGS samples**. For this, we first need to finish the filtering steps for the data according to **rovquantR**:

```{r}
## REMOVE SUSPECT SAMPLES ACCORDING TO HENRIK (THERE ARE NOT SUSPECT SAMPLES IN SWEDEN)
myFullData_NEW$alive$DNAID <- as.character(myFullData_NEW$alive$DNAID)
myFullData_NEW$alive <- myFullData_NEW$alive[!(myFullData_NEW$alive$DNAID %in% as.character(SUSPECT_NGS_SAMPLES$DNAID_RB)), ]
## REMOVE SUSPECT DEAD RECOVERIES ACCORDING TO HENRIK
myFullData_NEW$dead.recovery$DNAID <- as.character(myFullData_NEW$dead.recovery$DNAID)
myFullData_NEW$dead.recovery <- myFullData_NEW$dead.recovery[!(myFullData_NEW$dead.recovery$RovBaseId %in% as.character(SUSPECT_DeadRecoSAMPLES$Rovbase_ID)), ]

# ## REMOVE CUBS OF THE YEAR ACCORDING TO EVA HEDMARK (email 23/11/2025, document with explanation in AQEG Dropbox\AQEG Team Folder\RovQuant\DATA\RovbaseData\ROVBASE DOWNLOAD 20251121)
# myFullData_NEW$alive <- myFullData_NEW$alive[!(myFullData_NEW$alive$DNAID %in% c("D608410","D608411","D605997")), ]

## Remove individuals that died twice
myFullData_NEW$dead.recovery$Id <- as.character(myFullData_NEW$dead.recovery$Id)
IdDoubleDead <- myFullData_NEW$dead.recovery$Id[duplicated(myFullData_NEW$dead.recovery$Id)]
if(length(IdDoubleDead) > 0){
  duplicatedDeath <- NULL
  for(i in IdDoubleDead){
    tmp  <- which(myFullData_NEW$dead.recovery$Id == i & is.na(myFullData_NEW$dead.recovery$DeathCause_2))
    if(length(tmp)==0){tmp  <- which(myFullData_NEW$dead.recovery$Id == i)[-2]}##[CM] remove the second record.
    duplicatedDeath <- c(duplicatedDeath, tmp)
  }#i
  myFullData_NEW$dead.recovery <- myFullData_NEW$dead.recovery[-duplicatedDeath, ]
}#if

## Dead recoveries flagged by Henrik that should always be removed (email from the 18/12/2024)
myFullData_NEW$dead.recovery <- myFullData_NEW$dead.recovery[!myFullData_NEW$dead.recovery$RovBaseId %in% c("M495994", "M524051", "M524052", "M524053"), ]

## Remove pups killed before recruitment based on weight (cf. Henrik)
## 1) Remove individuals that are "Ja" in column "Doedt.individ..Unge" and recovered dead between March and November
myFullData_NEW$dead.recovery <- myFullData_NEW$dead.recovery[-which(myFullData_NEW$dead.recovery$Alder.pa.doedt.individ %in% "Unge" &
myFullData_NEW$dead.recovery$Month > 2 &
myFullData_NEW$dead.recovery$Month < 12), ]

## 2) Remove individuals that have a weight >0 and <4 between March and November format the weight correctly
myFullData_NEW$dead.recovery$Helvekt <- as.character(myFullData_NEW$dead.recovery$Helvekt)
myFullData_NEW$dead.recovery$Slaktevekt <- as.character(myFullData_NEW$dead.recovery$Slaktevekt)

##-- Convert to decimals
myFullData_NEW$dead.recovery$Helvekt <- as.numeric(gsub(",", ".", myFullData_NEW$dead.recovery$Helvekt))
myFullData_NEW$dead.recovery$Slaktevekt <- as.numeric(gsub(",", ".", myFullData_NEW$dead.recovery$Slaktevekt))

##-- Get the two weight columns together.
myFullData_NEW$dead.recovery$weight <- ifelse(
  !is.na(myFullData_NEW$dead.recovery$Helvekt),
  myFullData_NEW$dead.recovery$Helvekt,
  myFullData_NEW$dead.recovery$Slaktevekt)

##-- Assign negative values to nas to avoid issues
myFullData_NEW$dead.recovery$weight[is.na(myFullData_NEW$dead.recovery$weight)] <- -999
##-- check how many dead reco we remove and remove if more than 0
if(sum(myFullData_NEW$dead.recovery$weight > 0 &
       myFullData_NEW$dead.recovery$weight < 4 &
       myFullData_NEW$dead.recovery$Month < 12 &
       myFullData_NEW$dead.recovery$Month > 2)>0){
  myFullData_NEW$dead.recovery <- myFullData_NEW$dead.recovery[
    -which(myFullData_NEW$dead.recovery$weight > 0 &                                      myFullData_NEW$dead.recovery$weight < 4 &                                      myFullData_NEW$dead.recovery$Month < 12 &                                      myFullData_NEW$dead.recovery$Month > 2), ]
}
##-- Check how many dead reco with a weight of 0 kg and recovered between march and november
if(sum(myFullData_NEW$dead.recovery$Age %in% 0 &
       myFullData_NEW$dead.recovery$Month < 12 &
       myFullData_NEW$dead.recovery$Month > 2)>0){
  myFullData_NEW$dead.recovery[
    myFullData_NEW$dead.recovery$Age %in% 0 &
    myFullData_NEW$dead.recovery$Month < 12 &
    myFullData_NEW$dead.recovery$Month > 2,  ]
}

### ====    1.3. FILTER NGS & DEAD RECOVERY DATA ====
myFilteredData_NEW <- myFullData_NEW
## Subset to years of interest
myFilteredData_NEW$alive <- myFilteredData_NEW$alive[myFilteredData_NEW$alive$Year %in% years, ]
myFilteredData_NEW$dead.recovery <- myFilteredData_NEW$dead.recovery[myFilteredData_NEW$dead.recovery$Year %in% years, ]
## Subset to months of interest
myFilteredData_NEW$alive <- myFilteredData_NEW$alive[myFilteredData_NEW$alive$Month %in% unlist(sampling.months), ]

## Remove samples in Norrbotten in years not sampled
COUNTIESNorrbotten <- COUNTIES[COUNTIES$NAME_1 %in% "Norrbotten", ]
yearsSampledNorrb <- c(2016:2018,2023,2024)
is.Norr <- as.numeric(st_intersects(myFilteredData_NEW$alive, COUNTIESNorrbotten))
myFilteredData_NEW$alive <- myFilteredData_NEW$alive[- which(!myFilteredData_NEW$alive$Year %in% yearsSampledNorrb & !is.na(is.Norr)), ]

dim(myFilteredData_NEW$alive)
table(myFilteredData_NEW$alive$Year)
```

Another difference comes from the shapefile used to delineate the borders of Sweden. Using 'REGIONS' instead of 'COUNTRIES' leads to including more area and in particular a tiny part of Finland in the habitat.

```{r}
## DELINEATE A BUFFER AROUND ALL DETECTIONS 
myBufferedArea_NEW <- st_buffer( 
  myFilteredData_NEW$alive,                              
  dist = 60000 * 1.4) %>%   
  mutate(id = 1) %>% 
  group_by(id) %>%
  summarize() 

myStudyArea_NEW <- st_intersection(
  myBufferedArea_NEW,
  myStudyArea_NEW) %>% 
  group_by(id) %>% 
  summarize()

area_NEW <- st_area(myStudyArea_NEW)/1e6
area_NEW
```

```{r}
myBufferedArea <- st_buffer( 
  myFilteredData.sp$alive,
  dist = 60000 * 1.4) %>%
  mutate(id = 1) %>%
  group_by(id) %>% 
  summarize()

myStudyArea <- st_intersection(myBufferedArea,myStudyArea)
area_OLD <- st_area(myStudyArea)/1e6
area_OLD

##-- Difference between "new" and "old" study areas
area_diff <- area_NEW - area_OLD
```

In the end, the difference in study areas is `{r}area_diff` square kilometers.

```{r}
plot(st_geometry(GLOBALMAP))
plot(st_geometry(myStudyArea), border = "firebrick4", add = TRUE)
plot(st_geometry(myStudyArea_NEW), border = "navyblue", add = TRUE)
```

```{r}
myHabitat <- MakeHabitatFromRastersf( 
  poly = myStudyArea,
  habitat.r = habitatRasters[["Habitat"]],
  buffer = 60000,                               
  plot.check = F)
sum(myHabitat$habitat.r[]==1)

myHabitat_NEW <- MakeHabitatFromRastersf( 
  poly = myStudyArea_NEW,
  habitat.r = habitatRasters[["Habitat"]],
  buffer = 60000,                               
  plot.check = F)
sum(myHabitat_NEW$habitat.r[]==1)

```
