---
params:
  species: "bear"
  working_dir: "./data"
  data_dir: NULL
title: "`r paste(params$species, params$years[1], 'to', params$years[length(params$years)])`"
subtitle: "RovQuant - OPSCR Data Preparation"
output:
  html_document: 
    theme: "cerulean" 
    toc_depth: 2
    toc: yes
    toc_float: yes
    #css: bootstrap.css
    highlight: zenburn
  pdf_document: default
  word_document: default
---



```{r, enviro, echo = FALSE, warning = FALSE, message = FALSE}
years <- if(is.null(params$years)){ 
  2012:as.numeric(format(Sys.Date(), "%Y"))
} else { 
  params$years
}
n.years <- length(years)
samplingMonths <- params$samplingMonths
species <- params$species
data_dir <- params$data_dir
working_dir <- params$working_dir

if(sum(grep("bear", species, ignore.case = T))>0|sum(grep("bjorn", species, ignore.case = T))>0){
  engSpecies <- "bear"
  norSpecies <- "Bjørn"
}
if(sum(grep("wolf", species, ignore.case = T))>0|sum(grep("ulv", species, ignore.case = T))>0){
  engSpecies <- "wolf"
  norSpecies <- "Ulv"
}
if(sum(grep("wolverine", species, ignore.case = T))>0|sum(grep("jerv", species, ignore.case = T))>0){
  engSpecies <- "wolverine"
  norSpecies <- "Jerv"
}

months = c("January","February","March","April","May","June",
           "July","August","September","October","November","December")


##-- load libraries
library(kableExtra)
library(ggplot2)
library(sf)
library(mapview)
mapviewOptions(fgb = FALSE)


##-- Load pre-processed habitat shapefiles
data(COUNTRIESWaterHumans, envir = environment()) 

COUNTRIES <- COUNTRIESWaterHumans[COUNTRIESWaterHumans$ISO %in% c("SWE","NOR"), ] %>%
  dplyr::group_by(ISO) %>%
  dplyr::summarise()
```


### `r Sys.Date()`


# Introduction

This document summarizes the data preparation process of the **`r species`** DNA data available from [RovBase 3.0](https://rovbase.no/), for analysis with an OPSCR model as done by the [AQEG team](https://www.nmbu.no/en/projects/rovquant). In this particular instance, we focus on the **`r species`** data collected during the period **`r years[1]`** to **`r years[length(years)]`**.


------------------------------------------------------------------------

# Habitat definition
 
```{r, load habitat, echo = FALSE}
load(file.path(working_dir, "data/Habitat.RData"))
```



## Habitat characteristics

The habitat, sometimes called state-space (add REFS), describes the space available for the placement of individual activity centers. In the case of the `r species` OPSCR analysis, it is composed of 
`r  habitat$n.habwindows` habitat grid cells, each of size `r ( habitat$resolution/1000)^2` $km^2$.

```{r, hab grid, echo = FALSE, fig.align = "center"}
mapview( x = habitat$grid,
         hide = FALSE,
         legend = FALSE,
         map.types = c( "OpenTopoMap",
                        "Esri.WorldShadedRelief",
                        "OpenStreetMap.DE"))
```

## Habitat covariates 

```{r, hab covs1, echo = FALSE}
##-- Identify number of habitat covariates
covNames <- names(habitat$grid)[!names(habitat$grid) %in% c("Habitat","id", "x","y","geometry")]

##-- Plot layout
L <- length(covNames)
if(L < 6){ nrows <- 1 } else{
  if(L < 13){ nrows <- 2 } else {
    if(L < 22){ nrows <- 3 } else {
      if(L < 33){ nrows <- 4 } else {
        nrows <- 5
      }}}}
ncols <- ceiling(L/nrows)
```

In order to better capture the distribution of individual activity centers throughout the landscape, i.e. population density, spatial covariates can be used in the movement sub-model of the OPSCR.
In the case of the `r species` model, `r L` habitat covariates were included in the OPSCR model.

```{r, hab covs2, echo = FALSE}
par(mfrow = c(nrows,ncols))
for(c in 1:L){     
  plot( st_geometry(GLOBALMAP), 
        col = "gray80",
        main = covNames[c])
  plot(habitat$grid[ ,covNames[c]], add=T)
}#c
```

The prepared habitat characteristics are available in :
`r file.path(working_dir, "data/Habitat.RData")`



------------------------------------------------------------------------

# Detectors definition

```{r, det grid, echo = FALSE}
load(file.path(working_dir, "data/Detectors.RData"))
```



## Detectors characteristics

The surveyed area, also called detector-space (add REFS), describes the region that was searched for `r species` DNA samples. Although the area searched is continuous, it is discretized for practical reasons in the model. In the case of the `r species` OPSCR analysis, we used a grid with `r detectors$resolution/1000` km resolution. The area surveyed is determined by using a buffer around the DNA samples collected. 

It is composed of `r detectors$n.detectors` detector grid cells, each of size `r (detectors$resolution/1000)^2` $km^2$, leading to a total surveyed area of `r detectors$n.detectors*(detectors$resolution/1000)^2` $km^2$.

```{r, det grid, echo = FALSE, fig.align = "center"}
mapview(st_geometry(detectors$grid),
        hide = FALSE,
        legend = FALSE,
        map.types = c( "OpenTopoMap",
                       "Esri.WorldShadedRelief",
                       "OpenStreetMap.DE"))
```



## Detectors covariates
```{r, det covs1, echo = FALSE}
##-- Identify number of detector covariates
covNames <- names(detectors$grid)[!names(detectors$grid) %in% c("Habitat","id", "x","y","geometry")]

##-- Plot layout
L <- length(covNames)
if(L < 6){ nrows <- 1 } else{
  if(L < 13){ nrows <- 2 } else {
    if(L < 22){ nrows <- 3 } else {
      if(L < 33){ nrows <- 4 } else {
        nrows <- 5
      }}}}
ncols <- ceiling(L/nrows)
```

In order to better capture the variation in detection probability throughout the landscape as a result of e.g. a variation in search effort, spatial covariates can be used in the detection sub-model of the OPSCR.
In the case of the `r species` model, `r L` detector covariates were included in the OPSCR model.

```{r, det covs2, echo = FALSE}
par(mfrow = c(nrows,ncols))
for(c in 1:L){     
  plot( st_geometry(GLOBALMAP), 
        col = "gray80",
        main = covNames[c])
  plot(detectors$grid[ ,covNames[c]], border = NA, add=T)
}#c
```

------------------------------------------------------------------------

# Individual detection histories


```{r, det data, echo = FALSE}
##-- Load the most recent data from RovBase
myFullData.sp <- readMostRecent( 
  path = file.path(working_dir,"data"),
  pattern = paste0("Data_",species),
  extension = ".RData")

##-- list directories containing input files
dirNames <- list.dirs(file.path(working_dir, "nimbleInFiles"), full.names = F, recursive = F)

##-- path to inupt files
dirList <- paste0(working_dir, "/nimbleInFiles/",dirNames)

##-- Load different input files in a list
inputs <- lapply( dirList,
                 function(x){
                   load(list.files(x, full.names = T)[1])
                 })
names(inputs) <- dirNames
```



We can also add maps of the NGS samples collected year:

```{r NGS maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}
# alive.sf <- st_as_sf(alive)
# dead.sf <- st_as_sf(dead.recovery)
# COUNTRIES.sf <- st_as_sf(COUNTRIES)
  ##-- Plot NGS data
  par(mfrow = c(3,5), mar = c(0,0,2,0))
  for(t in 1:n.years){
    plot(st_geometry(COUNTRIES), col = "gray80")
    mtext(years[t], side = 1, line = -4)
    if(t == 3){ mtext("NGS data", side = 3, cex = 3, line = -2)}
    plot(st_geometry(myFullData.sp$alive[myFullData.sp$alive$Year %in% years[t], ]),
         add = TRUE, pch = 3, cex = 0.2, col = "navyblue")
  }#t


##-- NGS map
numRows <- ceiling(length(years)/5)
numCols <- 5
NGS_map <- ggplot(data = alive) +
  geom_sf(data = COUNTRIES,
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) +
  geom_sf(color = "black",
          alpha = 0.3, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
NGS_map

##-- Save maps as .png
grDevices::png(filename = file.path(dir.out, 
                         paste0(species, "_NGS_", years[1]," to ", years[length(years)], ".png")),
    width = 8, height = 6, units = "in", res = 300)
NGS_map
graphics.off()
```

and a series of maps for the dead recoveries each year:

```{r, Dead recovery maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}

##-- Plot Dead Recoveries data
par(mfrow = c(2,5), mar = c(0,0,2,0))
for(t in 1:n.years){
  plot(st_geometry(COUNTRIES), col = "gray80")
  mtext(years[t], side = 1, line = -4)
  if(t == 3){ mtext("Dead Recoveries", side = 3, cex = 3, line = -2)}
  plot(st_geometry(myFullData.sp$dead.recovery[myFullData.sp$dead.recovery$Year %in% years[t], ]),
       add = TRUE, pch = 3, cex = 0.2, col = "red")
}#t


dead_map <- ggplot(data = dead.recovery) +
  geom_sf(data = COUNTRIES, 
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) + 
  geom_sf(color = "black", alpha = 0.5, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
dead_map

##-- Save maps as .png
grDevices::png(filename = file.path(dir.out, 
                         paste0(engSpecies, "_DEAD_", years[1]," to ", years[length(years)], ".png")),
    width = 8, height = 6, units = "in", res = 300)
dead_map
graphics.off()
```



We can also look at some summary statistics based on individual detections, such as number of detections per individual (total), number of individual detections per individual (each year), pairwise distance between detections (in the same year ==> equivalent to sigma), distance between consecutive naive activity centers (equivalent to Tau.)



```{r data summary, echo =  F, fig.cap= "A) Nombre de détections par lièvre variable, par an, et par sexe, et B) distribution des distances entre deux detections du même individu au cours de la même année sur le site de Réotier. Les lignes pointillées montrent les valeurs moyennes pour chaque sexe.", }
##-- Histogram of numbers of detections per individual per year
df <- ngs %>%
  st_drop_geometry() %>%
  summarize(., 
            freq = n(),
            .by = c(id,sex,year))
                 
mu <- summarize( df, 
                 grp.mean = mean(freq),
                 .by = sex)

plot1 <- ggplot(df, aes(x = freq, fill = sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.5, binwidth = 1)+
  geom_vline( data = mu,
              aes( xintercept = grp.mean, color = sex),
              linetype = "dashed") +
  labs( tag = "A",
        x = "Number of detections per year",
        y = "Number of individuals") +
  theme_classic( ) +
  theme(legend.position = 'none')


##-- Histogram of distances between detections in a year
meanDist <- summarize( detDist, 
                       grp.mean = mean(dist),
                       .by = sex)

plot2 <- ggplot(detDist, aes(x = dist, fill = sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.5, binwidth = 100) +
  geom_vline( data = meanDist,
              aes( xintercept = grp.mean, color = sex),
              linetype = "dashed") +
  labs( tag = "B",
        x = "Pairwise distance between detections (m)",
        y = "Number of individuals") +
  theme_classic()

plot1 + plot2


```



```{r, echo = F}
##-- Number of NGS samples
kable( samples,
       align = "lc",
       caption = "Number of NGS samples per year and country") %>%
  kable_styling(full_width = F)

##-- Number of dead recoveries
kable( deadSamples,
       align = "lc",
       caption = "Number of DNA samples from dead animals per year and country") %>%
  kable_styling(full_width = F)
```




------------------------------------------------------------------------

# Local evaluation

```{r, save data, echo = F}
fileName <-  paste0("Data_", engSpecies, "_",params$modDate,".RData")

save(alive, 
     dead.recovery,
     IdDoubleSex,
     file = file.path(dir.out, fileName))
```

Finally, we save the cleaned **alive** and **dead.recovery** sf objects as a .RData file with name **`r fileName`** located in the `r engSpecies`-specific folder (`r dir.out`).




------------------------------------------------------------------------

# Metadata

```{r, metadata, echo = F}
utils::sessionInfo()
```
