---
params:
  species: "Brown bear"
  years: !r 2012:as.numeric(format(Sys.Date(), "%Y"))
  sex: c("Hann","Hunn")
  date: !r Sys.Date()   
  working_dir: !r getwd()
title: "`r paste(params$species, params$years[1], 'to', params$years[length(params$years)])`"
subtitle: "RovQuant - OPSCR Data Preparation"
output:
  html_document: 
    theme: "cerulean" 
    toc_depth: 2
    toc: yes
    toc_float: yes
    df_print: kable
    #css: bootstrap.css
    highlight: zenburn
  pdf_document: default
  word_document: default
---

```{r, enviro, echo = FALSE, warning = FALSE, message = FALSE}
years <- if(is.null(params$years)){ 
  2012:as.numeric(format(Sys.Date(), "%Y"))
} else { 
  params$years
}
date <- params$date
n.years <- length(years)
species <- params$species
working_dir <- params$working_dir

if(sum(grep("bear", species, ignore.case = T))>0|sum(grep("bjorn", species, ignore.case = T))>0){
  engSpecies <- "bear"
  norSpecies <- "Bjørn"
}
if(sum(grep("wolf", species, ignore.case = T))>0|sum(grep("ulv", species, ignore.case = T))>0){
  engSpecies <- "wolf"
  norSpecies <- "Ulv"
}
if(sum(grep("wolverine", species, ignore.case = T))>0|sum(grep("jerv", species, ignore.case = T))>0){
  engSpecies <- "wolverine"
  norSpecies <- "Jerv"
}

months = c("January","February","March","April","May","June",
           "July","August","September","October","November","December")


##-- load libraries
library(kableExtra)
library(ggplot2)
library(sf)
library(ggmap)
library(patchwork)
library(mapview)
mapviewOptions(fgb = FALSE)
library(rovquantR)

##-- Load pre-processed habitat shapefiles
data(COUNTRIESWaterHumans, envir = environment()) 

COUNTRIES <- COUNTRIESWaterHumans[COUNTRIESWaterHumans$ISO %in% c("SWE","NOR"), ] %>%
  dplyr::group_by(ISO) %>%
  dplyr::summarise()
```

### `r Sys.Date()`

# Introduction

This document summarizes the data preparation process of the **`r species`** DNA data available from [RovBase 3.0](https://rovbase.no/), for analysis with an OPSCR model as done by the [AQEG team](https://www.nmbu.no/en/projects/rovquant). In this particular instance, we focus on the **`r species`** data collected during the period **`r years[1]`** to **`r years[length(years)]`**.

------------------------------------------------------------------------

# Rovbase data

```{r, data, echo = FALSE}
load(file.path( working_dir, paste0("data/FilteredData_", engSpecies, "_", date, ".RData")))
```

## NGS data

```{r, ngs data, echo = FALSE, collapse = TRUE}

ngs <- data.alive$myData.sp

##-- Number of NGS samples collected per sex and per year 
numDet_sexYear <- table(ngs$Sex,ngs$Year)
if(length(dim(numDet_sexYear)) > 1){
  numDet_sexYear <- cbind(numDet_sexYear,rowSums(numDet_sexYear))
  colnames(numDet_sexYear) <- c(years,"total")
  numDet_sexYear <- rbind(numDet_sexYear,colSums(numDet_sexYear))
  rownames(numDet_sexYear) <- c(sex,"total")
}


##-- Number of individuals detected per sex and per year
numId_sexYear <- apply(table(ngs$Id,ngs$Sex,ngs$Year) > 0, c(2,3), sum)
if(length(dim(numId_sexYear)) > 1){
  numId_sexYear <- cbind(numId_sexYear,colSums(table(ngs$Id,ngs$Sex) > 0))
  colnames(numId_sexYear) <- c(years,"total")
  numId_sexYear <- rbind(numId_sexYear,colSums(numId_sexYear))
  rownames(numId_sexYear) <- c(sex,"total")
}


##-- Number of NGS samples per individual per year
df <- ngs %>%
  st_drop_geometry() %>%
  summarize(., 
            freq = n(),
            .by = c(Id,Sex,Year))
                 
mu <- summarize( df, 
                 grp.mean = mean(freq),
                 .by = Sex)


##-- Distance between detections per id and year 
IDs <- unique(ngs$Id)
detDist <- list()
for(i in IDs){
  tmp1 <- ngs[ngs$Id %in% i, ]
  tmp.list <- list()
  for(t in 1:n.years){
    tmp <- tmp1[tmp1$Year %in% years[t], ]
    if(nrow(tmp)>1){
          tmp.list[[t]] <- as.numeric(st_distance(tmp))
    }
  }#t
  thisDist <- unlist(tmp.list)
  detDist[[i]] <- cbind.data.frame(
    id = rep(i,length(thisDist)),
    sex = rep(unique(tmp1$Sex),length(thisDist)),
    dist = thisDist)
}#i
detDist <- do.call(rbind,detDist)
```

After filtering the available data to the area and period of interest (`r years[1]` to `r years[length(years)]`), we are left with the following number of non-invasive genetic samples:

```{r, ngs data1, echo = FALSE, collapse = TRUE}
numDet_sexYear
```

These samples led to the identification of `r numId_sexYear["total","total"]` individuals:

```{r, ngs data2, echo = FALSE, collapse = TRUE}
numId_sexYear
```

We can also look at the spatial distribution of these samples and individuals:

```{r, ngs data3, echo = FALSE, collapse = TRUE}
##-- Number of individuals detected 
ids <- apply(table(data.alive$myData.sp$Country_sample,
                   data.alive$myData.sp$Year,
                   data.alive$myData.sp$Id),
             c(1,2),
             function(x)sum(x>0))
ids <- rbind(ids,
             "Total" = apply(table(data.alive$myData.sp$Year,
                                   data.alive$myData.sp$Id),
                             1,
                             function(x)sum(x>0)))
ids <- cbind(ids,
             "Total" = c(apply(table(data.alive$myData.sp$Country_sample,
                                     data.alive$myData.sp$Id),
                               1,
                               function(x)sum(x>0)),
                         length(unique(data.alive$myData.sp$Id))))
ids
```

```{r NGS maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}
##-- NGS map
numRows <- ceiling(length(years)/5)
numCols <- 5
NGS_map <- ggplot(data = data.alive$myData.sp) +
  geom_sf(data = COUNTRIES,
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) +
  geom_sf(color = "black",
          alpha = 0.3, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
NGS_map
```

```{r, ngs data4, echo = FALSE, collapse = TRUE,fig.height = 6, fig.width = 12}
##-- Histogram of number of samples per individual per year
plot1 <- ggplot(df, aes(x = freq, fill = Sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.9, binwidth = 1)+
  geom_vline( data = mu,
              aes( xintercept = grp.mean, color = Sex),
              linetype = "dashed") +
  labs( tag = "A",
        x = "Number of NGS samples, per individual, per year",
        y = "Number of individuals") +
  theme_classic( ) +
  theme(legend.position = 'none')


##-- Histogram of distances between detections in a year
meanDist <- summarize( detDist, 
                       grp.mean = mean(dist),
                       .by = sex)

plot2 <- ggplot(detDist, aes(x = dist, fill = sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.9, binwidth = 1000) +
  geom_vline( data = meanDist,
              aes( xintercept = grp.mean, color = sex),
              linetype = "dashed") +
  labs( tag = "B",
        x = "Pairwise distance between detections (m)",
        y = "Number of individuals") +
  theme_classic()

##-- plots side by side
plot1 + plot2
```

## Dead recovery data

```{r, dead data1, echo = FALSE, collapse = TRUE}
ngs <- data.dead

##-- Number of samples collected per sex and per year 
numDet_sexYear <- table(ngs$Sex,ngs$Year)
if(length(dim(numDet_sexYear)) > 1){
  numDet_sexYear <- cbind(numDet_sexYear,rowSums(numDet_sexYear))
  colnames(numDet_sexYear) <- c(years,"total")
  numDet_sexYear <- rbind(numDet_sexYear,colSums(numDet_sexYear))
  rownames(numDet_sexYear) <- c(sex,"total")
}
numDet_sexYear
```

We can also plot the distribbution of those dead recoveries each year:

```{r, Dead recovery maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}
dead_map <- ggplot(data = data.dead) +
  geom_sf(data = COUNTRIES, 
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) + 
  geom_sf(color = "black", alpha = 0.5, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
dead_map
```

------------------------------------------------------------------------

# Habitat definition

```{r, load habitat, echo = FALSE}
load(file.path(working_dir, "data/Habitat.RData"))
```

## Habitat characteristics

The habitat, sometimes called state-space (add REFS), describes the space available for the placement of individual activity centers. In the case of the `r species` OPSCR analysis, it is composed of `r  habitat$n.habwindows` habitat grid cells, each of size `r ( habitat$resolution/1000)^2` $km^2$.

```{r, hab grid, echo = FALSE, fig.align = "center"}
mapview( x = habitat$grid,
         hide = FALSE,
         legend = FALSE,
         map.types = c( "OpenTopoMap",
                        "Esri.WorldShadedRelief",
                        "OpenStreetMap.DE"))
```

## Habitat covariates

```{r, hab covs1, echo = FALSE}
##-- Identify number of habitat covariates
covNames <- names(habitat$grid)[!names(habitat$grid) %in% c("Habitat","id", "x","y","geometry")]

##-- Plot layout
L <- length(covNames)
if(L < 6){ nrows <- 1 } else{
  if(L < 13){ nrows <- 2 } else {
    if(L < 22){ nrows <- 3 } else {
      if(L < 33){ nrows <- 4 } else {
        nrows <- 5
      }}}}
ncols <- ceiling(L/nrows)
```

In order to better capture the distribution of individual activity centers throughout the landscape, i.e. population density, spatial covariates can be used in the movement sub-model of the OPSCR. In the case of the `r species` model, `r L` habitat covariates were included in the OPSCR model.

```{r, hab covs2, echo = FALSE}
par(mfrow = c(nrows,ncols))
for(c in 1:L){     
  plot( st_geometry(GLOBALMAP), 
        col = "gray80",
        main = covNames[c])
  plot(habitat$grid[ ,covNames[c]], add=T)
}#c
```

The prepared habitat characteristics are available in : `r file.path(working_dir, "data/Habitat.RData")`

------------------------------------------------------------------------

# Detectors definition

```{r, det grid, echo = FALSE}
load(file.path(working_dir, "data/Detectors.RData"))
```

## Detectors characteristics

The surveyed area, also called detector-space (add REFS), describes the region that was searched for `r species` DNA samples. Although the area searched is continuous, it is discretized for practical reasons in the model. In the case of the `r species` OPSCR analysis, we used a grid with `r detectors$resolution/1000` km resolution. The area surveyed is determined by using a buffer around the DNA samples collected.

It is composed of `r detectors$n.detectors` detector grid cells, each of size `r (detectors$resolution/1000)^2` $km^2$, leading to a total surveyed area of `r detectors$n.detectors*(detectors$resolution/1000)^2` $km^2$.

```{r, det grid2, echo = FALSE, fig.align = "center"}
mapview(st_geometry(detectors$grid),
        hide = FALSE,
        legend = FALSE,
        map.types = c( "OpenTopoMap",
                       "Esri.WorldShadedRelief",
                       "OpenStreetMap.DE"))
```

## Detectors covariates

```{r, det covs1, echo = FALSE}
##-- Identify number of detector covariates
covNames <- names(detectors$grid)[!names(detectors$grid) %in% c("Habitat","id", "x","y","geometry")]

##-- Plot layout
L <- length(covNames)
if(L < 6){ nrows <- 1 } else{
  if(L < 13){ nrows <- 2 } else {
    if(L < 22){ nrows <- 3 } else {
      if(L < 33){ nrows <- 4 } else {
        nrows <- 5
      }}}}
ncols <- ceiling(L/nrows)
```

In order to better capture the variation in detection probability throughout the landscape as a result of e.g. a variation in search effort, spatial covariates can be used in the detection sub-model of the OPSCR. In the case of the `r species` model, `r L` detector covariates were included in the OPSCR model.

```{r, det covs2, echo = FALSE}
par(mfrow = c(nrows,ncols))
for(c in 1:L){     
  plot( st_geometry(GLOBALMAP), 
        col = "gray80",
        main = covNames[c])
  plot(detectors$grid[ ,covNames[c]], border = NA, add=T)
}#c
```

------------------------------------------------------------------------

# Individual detection histories

```{r, det data, echo = FALSE}
##-- list directories containing input files
dirNames <- list.dirs(file.path(working_dir, "nimbleInFiles"), full.names = F, recursive = F)

##-- path to inupt files
dirList <- paste0(working_dir, "/nimbleInFiles/",dirNames)

##-- Load different input files in a list
inputs <- lapply( dirList,
                 function(x){
                   load(list.files(x, full.names = T)[1])
                 })
names(inputs) <- dirNames
```

ONce formatted into individual detection histories, we can look at some summary statistics such as the number of detectors at which an individual is detected  (total), number of individual detections per individual (each year), pairwise distance between detections (in the same year ==\> equivalent to sigma), distance between consecutive naive activity centers (equivalent to Tau.)



**[...more coming soon...]**



```{r, eval = F}
##-- Number of NGS samples
kable( samples,
       align = "lc",
       caption = "Number of NGS samples per year and country") %>%
  kable_styling(full_width = F)

##-- Number of dead recoveries
kable( deadSamples,
       align = "lc",
       caption = "Number of DNA samples from dead animals per year and country") %>%
  kable_styling(full_width = F)
```

------------------------------------------------------------------------

# Local evaluation


**[...more coming soon...]**

------------------------------------------------------------------------

# Metadata

```{r, metadata, echo = F}
utils::sessionInfo()
```
