---
params:
  species: "bear"
  working_dir: "./data"
  data_dir: NULL
title: "`r paste(params$species, params$years[1], 'to', params$years[length(params$years)])`"
subtitle: "RovQuant - OPSCR Data Preparation"
output:
  html_document: 
    theme: "cerulean" 
    toc_depth: 2
    toc: yes
    toc_float: yes
    #css: bootstrap.css
    highlight: zenburn
  pdf_document: default
  word_document: default
---



```{r, enviro, echo = FALSE, warning = FALSE, message = FALSE}
years <- if(is.null(params$years)){ 
  2012:as.numeric(format(Sys.Date(), "%Y"))
} else { 
  params$years
}
n.years <- length(years)
samplingMonths <- params$samplingMonths
species <- params$species
data_dir <- params$data_dir
working_dir <- params$working_dir

if(sum(grep("bear", species, ignore.case = T))>0|sum(grep("bjorn", species, ignore.case = T))>0){
  engSpecies <- "bear"
  norSpecies <- "Bjørn"
}
if(sum(grep("wolf", species, ignore.case = T))>0|sum(grep("ulv", species, ignore.case = T))>0){
  engSpecies <- "wolf"
  norSpecies <- "Ulv"
}
if(sum(grep("wolverine", species, ignore.case = T))>0|sum(grep("jerv", species, ignore.case = T))>0){
  engSpecies <- "wolverine"
  norSpecies <- "Jerv"
}

months = c("January","February","March","April","May","June",
           "July","August","September","October","November","December")


##-- load libraries
library(kableExtra)
library(ggplot2)
library(sf)
library(mapview)
mapviewOptions(fgb = FALSE)


##-- Load age look-up table
# load(file.path(dir.in,"MISC DATA/age.lookup.table.RData"))

##-- Load pre-processed habitat shapefiles
#load(system.file("extdata", "Habitat_shp.RData", package = "rovquantR"))
data(COUNTRIESWaterHumans, envir = environment()) 

COUNTRIES <- COUNTRIESWaterHumans[COUNTRIESWaterHumans$ISO %in% c("SWE","NOR"), ] %>%
  dplyr::group_by(ISO) %>%
  dplyr::summarise()
```


### `r Sys.Date()`


# Introduction

This document summarizes the data preparation process of the **`r species`** DNA data available from [RovBase 3.0](https://rovbase.no/), for analysis with an OPSCR model as done by the [AQEG team](https://www.nmbu.no/en/projects/rovquant). In this particular instance, we focus on the **`r species`** data collected during the period **`r years[1]`** to **`r years[length(years)]`**.


------------------------------------------------------------------------

# Habitat definition
 
```{r, load habitat, echo = FALSE}
load(file.path(working_dir, "data/Habitat.RData"))
```



## Habitat characteristics

The habitat, sometimes called state-space (add REFS), describes the space available for the placement of individual activity centers. In the case of the `r species` OPSCR analysis, it is composed of 
`r  habitat$n.habwindows` habitat grid cells, each of size `r ( habitat$resolution/1000)^2` $km^2$.

```{r, hab grid, echo = FALSE}
mapview( x = habitat$buffered.habitat.poly,
         col.regions = c(NA,"blue","red"),
         hide = FALSE,
         legend = FALSE,
         map.types = c( "OpenTopoMap",
                        "Esri.WorldShadedRelief",
                        "OpenStreetMap.DE"))
```

## Habitat covariates 

```{r, hab covs}
##-- roads
plot( st_geometry(GLOBALMAP), 
      col = "gray80",
      main = "Distance to roads")
plot( st_geometry(studyArea),
      col = rgb(34/250, 139/250, 34/250,
                alpha = 0.5),
      add = T)
raster::plot(DistAllRoads,add=T)
# plot(st_geometry(detectors$main.detector.sp),
#      cex = scale(detRoads)/2,
#      pch = 16,
#      add = T)

```

```{r}

##-- Plot SkandObs and RovbaseObs
par(mfrow = c(2,2), mar = c(0,0,2,0))
for(t in 1:n.years){
  ## Rovbase + SkandObs smoothed
  plot(st_geometry(COUNTRIES), main = paste(years[t],"\n SkandObs + Rovbase smoothed binary"))
  #plot(ds.brickCont[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  plot(st_geometry(COUNTRIES),add=T)
}#t


for(t in 1:n.years){
  ## NGS DETECTIONS TOTAL
  tempTotal <- myFullData.sp$alive[myFullData.sp$alive$Year == years[t], ]
  NGS_TabTotal <- table(tempTotal$Country_sample)
  ID_TabTotal <- apply(table(tempTotal$Id, tempTotal$Country_sample), 2, function(x) sum(x>0))
  ## ALIVE DETECTIONS INSIDE STUDY AREA/SAMPLING PERIOD
  tempIn <- myFullData.sp$alive[myFullData.sp$alive$Year == years[t], ]
  NGS_TabIn <- table(tempIn$Country_sample)
  ID_TabIn <- apply(table(tempIn$Id, tempIn$Country_sample), 2, function(x) sum(x>0))
  ## PLOT NGS SAMPLES
  plot(st_geometry(GLOBALMAP), col="gray80")
  plot(st_geometry(studyArea), col = rgb(34/250, 139/250, 34/250, alpha = 0.5), add=T)
  # points(tempTotal, pch = 21, bg = "darkred")
  plot(st_geometry(tempIn), pch = 21, bg = "blue",add=T)
  ## ADD NUMBER OF NGS samples and IDs per COUNTRY
  graphics::text(x = 100000, y = 7200000, labels = paste(NGS_TabTotal[names(NGS_TabTotal)=="N"],"NGS"), cex = 1.1, col = "firebrick3", font = 2)
  graphics::text(x = 100000, y = 7270000, labels = paste(ID_TabTotal[names(NGS_TabTotal)=="N"], "IDs"), cex = 1.1, col = "firebrick3", font = 2)
  graphics::text(x = 820000, y = 6780000, labels = paste(NGS_TabTotal[names(NGS_TabTotal)=="S"],"NGS"), cex = 1.1, col = "navyblue", font = 2)
  graphics::text(x = 820000, y = 6850000, labels = paste(ID_TabTotal[names(NGS_TabTotal)=="S"], "IDs"), cex = 1.1, col = "navyblue", font = 2)
  ## ADD OVERALL NUMBERS
  mtext(text = years[t], side = 3, line = 1, cex = 1.5, font = 2)
  mtext(text = paste(sum(NGS_TabIn), "NGS/", sum(ID_TabIn), "IDs IN"), side = 3, line = 0)
  mtext(text = paste(sum(NGS_TabTotal)-sum(NGS_TabIn), "NGS/", sum(ID_TabTotal)-sum(ID_TabIn), "IDs OUT"), side = 3, line = -1)
}#t



  # ##-- plot check 
  # if(myVars$plot.check){
  #   pdf(file.path(WDFigures, "SamplesVsObs_years.pdf"),width = 12,height = 10)
  #   
  #   ##-- Plot NGS data
  #   par(mfrow = c(3,5), mar = c(0,0,2,0))
  #   for(t in 1:n.years){
  #     plot(st_geometry(COUNTRIES), col = "gray80")
  #     mtext(years[t], side = 1, line = -4)
  #     if(t == 3){ mtext("NGS data", side = 3, cex = 3, line = -2)}
  #     plot(st_geometry(myFullData.sp$alive[myFullData.sp$alive$Year %in% years[t], ]),
  #          add = TRUE, pch = 3, cex = 0.2, col = "navyblue")
  #   }#t
  #   
  #   ##-- Plot Dead Recoveries data
  #   par(mfrow = c(2,5), mar = c(0,0,2,0))
  #   for(t in 1:n.years){
  #     plot(st_geometry(COUNTRIES), col = "gray80")
  #     mtext(years[t], side = 1, line = -4)
  #     if(t == 3){ mtext("Dead Recoveries", side = 3, cex = 3, line = -2)}
  #     plot(st_geometry(myFullData.sp$dead.recovery[myFullData.sp$dead.recovery$Year %in% years[t], ]),
  #          add = TRUE, pch = 3, cex = 0.2, col = "red")
  #   }#t
  #   
  #   ##-- Plot SkandObs and RovbaseObs
  #   par(mfrow = c(2,3), mar = c(0,0,2,0))
  #   for(t in 1:n.years){
  #     ## SkandObs Binary
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n SkandObs binary"))
  #     plot(r.skandObsSamplesBinary[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES),add=T)
  #     
  #     ## Rovbase Binary
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n Rovbase binary"))
  #     plot(r.OtherSamplesBinary[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES),add=T)
  #     
  #     ## Rovbase + SkandObs Binary
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n SkandObs + Rovbase binary"))
  #     plot(r.SkandObsOtherSamplesBinary[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES), add=T)
  #     
  #     ## SkandObs continuous
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n SkandObs continuous"))
  #     plot(r.skandObsSamplesContinuous[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(COUNTRIES,add=T)
  #     
  #     ## Rovbase continuous
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n Rovbase continuous"))
  #     plot(r.OtherSamplesContinuous[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES),add=T)
  #     
  #     ## Rovbase + SkandObs smoothed
  #     plot(st_geometry(COUNTRIES), main = paste(years[t],"\n SkandObs + Rovbase smoothed binary"))
  #     plot(ds.brickCont[[t]], main=paste(year,"\n Rovbase binary"), box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES),add=T)
  #   }#t
  #   
  #   ##-- Plot Data and Obs side by side
  #   par(mfrow = c(1,3), mar = c(0,0,2,0))
  #   for(t in 1:n.years){
  #     ##-- NGS data
  #     plot(st_geometry(COUNTRIES), col = "gray80")
  #     mtext("NGS data", side = 1, line = -4)
  #     plot(st_geometry(myFullData.sp$alive[myFullData.sp$alive$Year %in% years[t], ]),
  #          add = TRUE, pch = 3, cex = 0.2, col = "navyblue")
  #     
  #     ##-- Dead recoveries data
  #     plot(st_geometry(COUNTRIES), col = "gray80")
  #     mtext("Dead recoveries", side = 1, line = -4)
  #     mtext(years[t], side = 3, cex = 3, line = -2)
  #     plot(st_geometry(myFullData.sp$dead.recovery[myFullData.sp$dead.recovery$Year %in% years[t], ]),
  #          add = TRUE, pch = 3, cex = 0.2, col = "red")
  #     
  #     ##-- Rovbase + SkandObs Binary
  #     plot(st_geometry(COUNTRIES), col = "gray80")
  #     mtext("SkandObs + Rovbase", side = 1, line = -4)
  #     plot(ds.brickCont[[t]], box=F, axes=F,add=T)
  #     plot(st_geometry(COUNTRIES),add=T)
  #   }#t
  #   
  #   
  #   for(t in 1:n.years){
  #     ## NGS DETECTIONS TOTAL
  #     tempTotal <- myFullData.sp$alive[myFullData.sp$alive$Year == years[t], ]
  #     NGS_TabTotal <- table(tempTotal$Country_sample)
  #     ID_TabTotal <- apply(table(tempTotal$Id, tempTotal$Country_sample), 2, function(x) sum(x>0))
  #     ## ALIVE DETECTIONS INSIDE STUDY AREA/SAMPLING PERIOD
  #     tempIn <- myFullData.sp$alive[myFullData.sp$alive$Year == years[t], ]
  #     NGS_TabIn <- table(tempIn$Country_sample)
  #     ID_TabIn <- apply(table(tempIn$Id, tempIn$Country_sample), 2, function(x) sum(x>0))
  #     ## PLOT NGS SAMPLES
  #     plot(st_geometry(GLOBALMAP), col="gray80")
  #     plot(st_geometry(studyArea), col = rgb(34/250, 139/250, 34/250, alpha = 0.5), add=T)
  #     # points(tempTotal, pch = 21, bg = "darkred")
  #     plot(st_geometry(tempIn), pch = 21, bg = "blue",add=T)
  #     ## ADD NUMBER OF NGS samples and IDs per COUNTRY
  #     graphics::text(x = 100000, y = 7200000, labels = paste(NGS_TabTotal[names(NGS_TabTotal)=="N"],"NGS"), cex = 1.1, col = "firebrick3", font = 2)
  #     graphics::text(x = 100000, y = 7270000, labels = paste(ID_TabTotal[names(NGS_TabTotal)=="N"], "IDs"), cex = 1.1, col = "firebrick3", font = 2)
  #     graphics::text(x = 820000, y = 6780000, labels = paste(NGS_TabTotal[names(NGS_TabTotal)=="S"],"NGS"), cex = 1.1, col = "navyblue", font = 2)
  #     graphics::text(x = 820000, y = 6850000, labels = paste(ID_TabTotal[names(NGS_TabTotal)=="S"], "IDs"), cex = 1.1, col = "navyblue", font = 2)
  #     ## ADD OVERALL NUMBERS
  #     mtext(text = years[t], side = 3, line = 1, cex = 1.5, font = 2)
  #     mtext(text = paste(sum(NGS_TabIn), "NGS/", sum(ID_TabIn), "IDs IN"), side = 3, line = 0)
  #     mtext(text = paste(sum(NGS_TabTotal)-sum(NGS_TabIn), "NGS/", sum(ID_TabTotal)-sum(ID_TabIn), "IDs OUT"), side = 3, line = -1)
  #   }#t
  #   graphics.off()
  # }
```


The prepared habitat characteristics are availale in :
`r file.path(working_dir, "data/Habitat.RData")`



------------------------------------------------------------------------

# Detectors definition

```{r, det grid, echo = FALSE}
load(file.path(working_dir, "data/Detectors.RData"))
```



## Detectors characteristics

The surveyed area, also called detector-space (add REFS), describes the region that was searched for `r species` DNA samples. Although the area searched is continuous, it is discretized for practical reasons in the model. In the case of the `r species` OPSCR analysis, we used a grid with `r detResolution/1000` km resolution. The area surveyed is determined by using a buffer around the DNA samples collected. 

It is composed of `r detectors$n.detectors` detector grid cells, each of size `r (detectors$resolution/1000)^2` $km^2$, leading to a total surveyed area of `r detectors$n.detectors*(detectors$resolution/1000)^2` $km^2$.

```{r, det grid, echo = FALSE, fig.align = "center"}
#| label: Final plot
#| echo: false 
#| warning: false
#| message: false
#| fig-cap: "Number of detections per lièvre variable and by sex, and B) distances between detections of the same individual in a given year in the Réotier nature reserve." 
#| out-width: 100%  
mapview( detectors$grid,
         burst = TRUE,
         map.types = c( "OpenTopoMap",
                        "Esri.WorldShadedRelief",
                        "OpenStreetMap.DE")) 

```



## Detectors covariates
```{r, detectors covariates}
par(mfrow = c(1,1))
myCol <- terrain.colors(nrow(COUNTIES))
plot( st_geometry(GLOBALMAP),
      col = "gray80", main = " Counties")
plot( st_geometry(studyArea),
      col = rgb(34/250, 139/250, 34/250, alpha = 0.5), add = T)
plot(st_geometry(detectors$main.detector.sp),
     col = myCol[detCounties], pch = 16, cex = 0.8, add = T)
```  


------------------------------------------------------------------------

# Individual detection histories

We can also add maps of the NGS samples collected year:

```{r NGS maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}
# alive.sf <- st_as_sf(alive)
# dead.sf <- st_as_sf(dead.recovery)
# COUNTRIES.sf <- st_as_sf(COUNTRIES)
  ##-- Plot NGS data
  par(mfrow = c(3,5), mar = c(0,0,2,0))
  for(t in 1:n.years){
    plot(st_geometry(COUNTRIES), col = "gray80")
    mtext(years[t], side = 1, line = -4)
    if(t == 3){ mtext("NGS data", side = 3, cex = 3, line = -2)}
    plot(st_geometry(myFullData.sp$alive[myFullData.sp$alive$Year %in% years[t], ]),
         add = TRUE, pch = 3, cex = 0.2, col = "navyblue")
  }#t


##-- NGS map
numRows <- ceiling(length(years)/5)
numCols <- 5
NGS_map <- ggplot(data = alive) +
  geom_sf(data = COUNTRIES,
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) +
  geom_sf(color = "black",
          alpha = 0.3, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
NGS_map

##-- Save maps as .png
grDevices::png(filename = file.path(dir.out, 
                         paste0(species, "_NGS_", years[1]," to ", years[length(years)], ".png")),
    width = 8, height = 6, units = "in", res = 300)
NGS_map
graphics.off()
```

and a series of maps for the dead recoveries each year:

```{r, Dead recovery maps, echo = F, fig.align = "center", fig.height = 8, fig.width = 12}

##-- Plot Dead Recoveries data
par(mfrow = c(2,5), mar = c(0,0,2,0))
for(t in 1:n.years){
  plot(st_geometry(COUNTRIES), col = "gray80")
  mtext(years[t], side = 1, line = -4)
  if(t == 3){ mtext("Dead Recoveries", side = 3, cex = 3, line = -2)}
  plot(st_geometry(myFullData.sp$dead.recovery[myFullData.sp$dead.recovery$Year %in% years[t], ]),
       add = TRUE, pch = 3, cex = 0.2, col = "red")
}#t


dead_map <- ggplot(data = dead.recovery) +
  geom_sf(data = COUNTRIES, 
          aes(fill = ISO),
          alpha = 0.3,
          color = NA) + 
  geom_sf(color = "black", alpha = 0.5, size = 0.8, pch = 3) +
  facet_wrap(~Year, nrow = numRows, ncol = numCols) +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_blank())
dead_map

##-- Save maps as .png
grDevices::png(filename = file.path(dir.out, 
                         paste0(engSpecies, "_DEAD_", years[1]," to ", years[length(years)], ".png")),
    width = 8, height = 6, units = "in", res = 300)
dead_map
graphics.off()
```


```{r, map quarto}
#| label: GPS track plot 
#| message: false 
#| warning: false
#| echo: false 
#| fig-cap: "Transects parcourus chaque année à la recherche de crottes de lièvres variables sur le site de Réotier." 
#| out-width: 120%  
colYears <- as.list(met.brewer( name = "Isfahan1",
                      n = length(searchTracks.inter), 
                      type="continuous"))
colOccasions <- hcl.colors(n = 5, palette = "Earth")

myPlots <- list()
for(t in 1:n.years){
  myPlots[[t]] <- ggplot(searchTracks.inter[[t]]) + 
    geom_sf(aes(col = as.factor(occasion)),
            show.legend = FALSE) +
    scale_color_manual(values = colOccasions) +
    theme( axis.line = element_blank(), 
           axis.text = element_blank(),
           axis.ticks = element_blank(),
           axis.title = element_blank()) +
     labs(title = years[t])
}#t

grid.arrange(arrangeGrob(myPlots[[1]],
                         myPlots[[2]],
                         myPlots[[3]],
                         myPlots[[4]],
                         myPlots[[5]],
                         myPlots[[6]],
                         myPlots[[7]],
                         myPlots[[8]],
                         myPlots[[9]],
                         myPlots[[10]],ncol = 2))

```

We can also look at some summary statistics based on individual detections, such as number of detections per individual (total), number of individual detections per individual (each year), pairwise distance between detections (in the same year ==> equivalent to sigma), distance between consecutive naive activity centers (equivalent to Tau.)



```{r data summary, echo =  F, fig.cap= "A) Nombre de détections par lièvre variable, par an, et par sexe, et B) distribution des distances entre deux detections du même individu au cours de la même année sur le site de Réotier. Les lignes pointillées montrent les valeurs moyennes pour chaque sexe.", }
##-- Histogram of numbers of detections per individual per year
df <- ngs %>%
  st_drop_geometry() %>%
  summarize(., 
            freq = n(),
            .by = c(id,sex,year))
                 
mu <- summarize( df, 
                 grp.mean = mean(freq),
                 .by = sex)

plot1 <- ggplot(df, aes(x = freq, fill = sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.5, binwidth = 1)+
  geom_vline( data = mu,
              aes( xintercept = grp.mean, color = sex),
              linetype = "dashed") +
  labs( tag = "A",
        x = "Number of detections per year",
        y = "Number of individuals") +
  theme_classic( ) +
  theme(legend.position = 'none')


##-- Histogram of distances between detections in a year
meanDist <- summarize( detDist, 
                       grp.mean = mean(dist),
                       .by = sex)

plot2 <- ggplot(detDist, aes(x = dist, fill = sex)) +
  geom_histogram( position = "dodge", 
                  alpha = 0.5, binwidth = 100) +
  geom_vline( data = meanDist,
              aes( xintercept = grp.mean, color = sex),
              linetype = "dashed") +
  labs( tag = "B",
        x = "Pairwise distance between detections (m)",
        y = "Number of individuals") +
  theme_classic()

plot1 + plot2


```



```{r, echo = F}
##-- Number of NGS samples
kable(samples, align = "lc",
      caption = "Number of NGS samples per year and country") %>%
  kable_styling(full_width = F)

##-- Number of dead recoveries
kable(deadSamples, align = "lc",
      caption = "Number of DNA samples from dead animals per year and country") %>% kable_styling(full_width = F)
```




------------------------------------------------------------------------

# Local evaluation

```{r, save data, echo = F}
fileName <-  paste0("Data_", engSpecies, "_",params$modDate,".RData")

save(alive, 
     dead.recovery,
     IdDoubleSex,
     file = file.path(dir.out, fileName))
```

Finally, we save the cleaned **alive** and **dead.recovery** sf objects as a .RData file with name **`r fileName`** located in the `r engSpecies`-specific folder (`r dir.out`).




------------------------------------------------------------------------

# Metadata

```{r, metadata, echo = F}
utils::sessionInfo()
```
